---
title: "Logistic regression"
subtitle: "Regression for Linguists"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
# footer: "Lecture 1.1 - R und RStudio"
lang: en
date: "`r Sys.Date()`"
format:
  html:
    number-sections: true
    number-depth: 3
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
bibliography: references.bib
csl: apa.csl
execute:
  eval: true
  echo: true
  message: false
  error: false
  warning: false
  fig-align: "center"
  fig-asp: .618
---

This lecture jumpts to chapter 12 'Genearlised Lienar Models 1: Logistic Regression" [@winter_statistics_2019]. We're skipping a few chapters, which I encourage you to go through on your own. However, they cover topics that you presumably have covered in previous courses (namely significance testing, *t*-values and *p*-values).

## Learning Objectives {.unnumbered .unlisted}

Today we will learn...


## Set-up environment  {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
```

We'll also need to load in our required packages. Hopefully you've already install the required packages (if not, go to @sec-software).

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               lme4,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra)
```

## Binomial distribution

## Generalised linear models

## Logistic regression

## Log-odds, odds ratio, and probabilities


## Summary

-   we saw that the equation for a straight line boils down to its intercept and slope

-   we fit our first linear model with a categorical predictor

### Important terms {.unnumbered .smaller}

```{r}
#| echo: false
tribble(
 ~"term", ~"description/other terms",
 
) %>% kable() %>% kable_styling()
```


## Learning Objectives {.unnumbered .unlisted}

Today we learned...


## Task



## Literaturverzeichnis {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


