---
title: "Logistic regression"
subtitle: "Regression for Linguists"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
# footer: "Lecture 1.1 - R und RStudio"
lang: en
date: "10/12/2023"
date-modified: last-modified
---

This lecture jumps to chapter 12 \'Genearlised Lienar Models 1: Logistic Regression' [@winter_statistics_2019]. We're skipping a few chapters, which I encourage you to go through on your own. However, they cover topics that you presumably have covered in previous courses (namely significance testing, *t*-values and *p*-values).

## Learning Objectives {.unnumbered .unlisted}

Today we will learn...

- how to model binomial data with logistic regression
- how to interpret log-odds and odds ratio

## Set-up environment  {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
options(pillar.sigfig = 5)
```

```{r}
library(broman)
# function to format p-values
format_pval <- function(x){
  if (x < .001) return(paste('<', '.001'))
  if (x < .01) return(paste('<', '.01'))
  if (x < .05) return(paste('<', '.05'))
  paste('=', myround(x, 3))  # if above .05, print p-value to 3 decimalp points
}
```


We'll also need to load in our required packages. Hopefully you've already install the required packages (if not, go to @sec-software).

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               lme4,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra,
               googlesheets4)

# tell googlesheets4 we don't want private
gs4_deauth()
```

```{r}
# set preferred ggplot2 theme
theme_set(theme_bw() + theme(plot.title = element_text(size = 10)))
```



## Generalised linear models

Logistic regression is a type of genearlised linear model (GLM), and is used to model binomial response data. Whereas continuous response variables, such as reaction times, assume a normal distribution (a.k.a., a Gaussian distribution), logistic regression assumes a binomial distribution (a.k.a., Bernoulli distribution). These are formalised in equations \ref{eq-normal}, where $\mu$ and $\sigma$ correspond to the mean and standard deviation, and \ref{eq-binomial}, where $N$ and $p$ refer to the number of trials and the probability of $y$ being $1$ or $0$. 

\begin{align}
y &\sim Normal(\mu,\sigma) \label{eq-normal} \\ 
y &\sim binomial(N = 1, p) \label{eq-binomial}
\end{align}  

Don't stress about this for now, I find the math behind everything will start to make more sense the more often you see it. However, *some* math is necessary in order to understand the output of our models, namely the relation between probabilities, odds, and log odds.

### Log-odds, odds ratio, and probabilities

In logistic regression, we the probability ($p$) of observing one outcome or another as a function of a predictor variable. In linguistic research, these outcomes could be the absence or presence of some phenomenon (pause, schwa, etc.) or button responses (yes/no, accept/reject). In logistic regression, we describe the probability, odds, or log-odds of a particular outcome over another.

```{r}
#| echo: false
df_odds <-
  tibble(
  log_odds = seq(from = -5, to = 5, by = 0.25),
  odds = exp(log_odds),
  prob = plogis(log_odds),
  q = 1 - prob
)
```

Probability is quite intuitive, and ranges from 0 (no chance) to 1 (certain). A 50% chance corresponds to a probability of 0.5. You're also likely familiar with odds, which can range from 0 to infinity. Odds are often used in betting, such as *the odds that I'll win are 2:1*, which corresponds to $\frac{2}{1} = 2$ in favour of my winning. Conversely, *the odds that you'll win are 1:2*, corresponding to  $\frac{1}{2} = 0.5$, meaning it's less likely that you'll win compared to you losing. If the odds are even, then: $\frac{1}{1} = 1$. So, odds of 1 correspond to a probability of 0.5. Log-odds are just the logarithmically-transformed odds: $log(2) =$ `r log(2)`; $log(0.5) =$ `r log(0.5)`; $log(1) =$ `r log(1)`. Probability can also be computed using the odds, as shown in \ref{eq-odds}: $\frac{2}{1+2} =$ `r 2/(1+2)`; $\frac{1}{1+1} =$ `r 1/(1+1)`; $\frac{0.5}{1+0.5} =$ `r 0.5/(1+0.5)`.

We can get the probability from a log odds value using `plogis()`, which performs the following calculation:

\begin{equation}
p = \frac{exp(log\;odds)}{1 + exp(log\;odds)} = \frac{odds}{1 + odds} \label{eq-odds}
\end{equation} 

@tbl-odds gives an example of how the three relate to each other. The grey cells are all where chances re 50/50, with increasingly more likely (green) or less likely (red) values/

```{r}
#| echo: false
#| label: tbl-odds
#| tbl-cap: Comparison of different values of probabilities/odds/log-odds
library(gt)
col_func <- colorRampPalette(c('lightblue', '#DE77AE'))

# Test function:

# col_func(4)

df_odds |> 
  filter(row_number() %% 5 == 1) |> 
  select(-q) |> 
  mutate(row = row_number()) %>%
  pivot_longer(
    cols = c(prob, odds, log_odds)
  ) |> 
  # pivot_wider(names_from = Cor, values_from = value)
  pivot_wider(
    # id_cols = names,
    names_from = row,
    values_from = value
  ) |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```

This relationship is demonstrated in @fig-odds. Take your time to really understand these plots, as it will help understand the output of our models.

```{r}
#| echo: false
#| label: fig-odds
#| fig-cap: Relationship between probability, odds, and log-odds
#| fig-asp: .35
fig_log_odds <-
  df_odds |> 
  ggplot() +
  aes(x = log_odds, y = odds) +
  geom_line() +
  scale_x_continuous(breaks = seq(-5, 5, by = 1)) +
  labs(title = "Odds versus log-odds") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "slategrey")

fig_prob_odds <-
  df_odds |> 
  ggplot() +
  aes(x = prob, y = odds) +
  geom_line() +
  scale_x_continuous(breaks = seq(-5, 5, by = 1)) +
  labs(title = "Odds versus probability") +
  geom_vline(xintercept = .5, linetype = "dashed", colour = "slategrey")

fig_prob_log <-
  df_odds |> 
  ggplot() +
  aes(x = prob, y = log_odds) +
  geom_line() +
  scale_x_continuous(breaks = seq(-5, 5, by = 1)) +
  labs(title = "Odds versus probability") +
  geom_vline(xintercept = .5, linetype = "dashed", colour = "slategrey")

fig_log_odds + fig_prob_odds + fig_prob_log + plot_annotation(tag_levels = "A")
```

## Logistic regression

I find the more we talk about the math behind the models before even running a model, the more overwhelmed we become. So, let's run our first logistic regression and then dissect it to understand it. Most relevant to the output of a logistic regression model is @fig-odds C, as the model will output log-odds, and we most likely want to interpret them in terms of probabilities.

We'll use a dataset from @biondo_yesterday_2022, an eye-tracking reading study exploring the processing of adverb-tense concord in Spanish past and future tenses. Participants read sentences that began with a temporal adverb (e.g., yesterday/tomorrow), and had a verb marked with the congruent or incongruent tense (past/future). We will look at the measure *regression in* at the *verb* region.

Let's start by loading in the data:

```{r}
df_tense <-
  read_csv(here("data", "Biondo.Soilemezidi.Mancini_dataset_ET.csv"),
           locale = locale(encoding = "Latin1") # for special characters in Spanish
           ) |> 
  mutate(gramm = ifelse(gramm == "0", "ungramm", "gramm")) |> 
  clean_names() |> 
  filter(roi == 4,
         adv_type == "Deic")
```

### EDA

And conducting a quick EDA: print summaries and plot the response variables.

```{r}
head(df_tense)
```

Let's look at only our binomial dependent variables, regression in (`ri`) and regression out (`ro`). For each variable, `1` indicates a regression in/out, `0` indicates there was no regression in/out.

```{r}
df_tense |> 
  select(roi, ri, ro) |> 
  summary()
```

Let's plot out our dependent variable of interest: regression in.

```{r}
# make grammaticality a factor
df_tense |> 
  mutate(gramm = as_factor(gramm)) 
```


```{r}
#| echo: false

facet_labels <-
  c(
    "ri" = "Reg. In",
    "ro" = "Reg. Out",
    "Future" = "Future",
    "Past" = "Past"
  )

fig_reg <-
  df_tense |> 
  filter(roi == "4") |> 
  # pivot_longer(
  #   cols = c(ri, ro),
  #   names_to = "measure",
  #   values_to = "regression"
  # ) |> 
  mutate(gramm = as_factor(gramm),
         ri = ifelse(ri == 1, "yes", "no")) |> 
  drop_na(ri) |> 
  ggplot() +
  labs(title = "Count of regressions in/out") +
  aes(x = gramm, fill = ri) +
  geom_bar() +
  facet_grid(.~verb_t, labeller = as_labeller(facet_labels)) 

fig_reg_prop <-
  df_tense |> 
  filter(roi == "4") |> 
  pivot_longer(
    cols = c(ri, ro),
    names_to = "measure",
    values_to = "regression"
  ) |> 
  mutate(gramm = as_factor(gramm),
         regression = as_factor(regression)) |> 
  drop_na(regression) |> 
  ggplot() +
  labs(title = "Proportion of regressions in/out") +
  aes(x = gramm, fill = regression) +
  geom_bar(position = "fill") +
  facet_grid(measure~verb_t) 

fig_reg
```

It looks like there are more regressions in for the grammatical versus ungrammatical conditions in both the future and past tenses. There doesn't seem to be a large difference between the two tenses in overall regressions in, nor in the effect of grammaticality on the proportion of regressions in. We can also see that it was more likely overall for there to *not* be a regression in (versus for there to be a regression in).

### Model

Now let's run our model. Verb tense and grammaticality are each two-level factors, so we'll want to set sum coding for each of them. Let's set `past` and `grammatical` to $-0.5$, and `future` and `ungrammatical` to `+0.5`.

```{r}
df_tense$verb_t <- as.factor(df_tense$verb_t)
levels(df_tense$verb_t)
contrasts(df_tense$verb_t) <- c(+0.5, -0.5)
contrasts(df_tense$verb_t)
```


```{r}
df_tense$gramm <- as.factor(df_tense$gramm)
levels(df_tense$gramm)
contrasts(df_tense$gramm) <- c(-0.5, +0.5)
contrasts(df_tense$gramm)
```

Now that we've set our contrasts (if you have continuous predictors, you would centre and potentially standardise them instead), we can fit our model. We use the `glm()` function to fit a genearlised linear model, and use the argument `family = "binomial"` to indicate our data are binomial.

```{r}
fit_tense_ri <-
  glm(ri ~ verb_t*gramm,
    data = df_tense,
    family = "binomial")
```

What do our coefficients look like?

```{r}
tidy(fit_tense_ri) %>%
  mutate(p.value = as.numeric(p.value)) |> 
  mutate(p.value = round(p.value*4,10)
         ) |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```

Let's first consider the estimates. The intercept is negative, meaning it is below 0. Verb tense is positive, meaning that there are more regressions in for the future compared to the past, holding grammaticality constant. Grammaticality is positive, meaning that there were more regressions in for the ungrammatical than grammatical conditions. But what does zero mean here? Logistic regression gives the estimates in log-odds. This means that a value of 0 means there is an equal probability of a regression in or out for both conditions (as in @tab-odds), i.e., the slope is flat (or not significantly different from 0). How can we convert our log-odds estimates to something more interpretable, like probabilities? Recall the equation in \ref{eq-odds}, which would require a lot of typing. Luckily, we can just use the `plogis()` function, which takes a log-odds value and spits out the corresponding probability. We can also just use the `exp()` function to get the odds ratio from the log-odds.

```{r}
plogis(-1.23) # intercept prob
plogis(0.0277) # tense prob

exp(-1.23) # intercept odds
exp(0.0277) # tense odds
```

This is great, but a bit tedious. We can also just feed a tibble column through the `plogis()` and `exp()` functions to print a table with the relevant probabilities and odds. 

```{r}
tidy(fit_tense_ri) %>%
  mutate(p.value = round(p.value*4,10),
         prob = plogis(estimate),
         odds = exp(estimate)
         ) |> 
  mutate_if(is.numeric, round, 4) |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```

We see that the odds of the future tense have a regression in versus the past tense is ~1, with the corresponding probability of 0.51. Unsuprisingly, we see this *p*-value indicates this effect was not significant (*p* > .05), and the *z*-value (note: not *t*-value!) is also low.


::: {.callout-note}

### *z*-values

*z*-values correspond to the estimate divided by the standard error. It's interpretation is similar to that of the *t*-value: a *z*-value of ~2 or higher will likely have a *p*-value below 0.05. 
:::

The interaction term is negative, what does this mean? We can interpret this as indicating that the effect of congruence is different in either level of tense. These effects are often more easily interpreted with a visualisation, e.g., using the `plot_model()` function from the `sjPlot` package. This effect is not significant, however.

```{r}
#| label: fig-inter
#| fig-cap: "Interaction plot of"
sjPlot::plot_model(fit_tense_ri, 
                   type = "eff", 
                   terms = c("gramm", "verb_t")) + 
  geom_line(position = position_dodge(0.1))
```

We can also use the `predict()` function to extract the predicted values for each condition. We could just simply print the predicted values (`predict(fit_tense_ri)`), append the predicted values to the data frame

```{r}
# make sure dataset is the same length as the model data
df_tense_v <-
  df_tense |> 
  filter(roi == "4") |> 
  drop_na(ri)

# append model estimates
augment(fit_tense_ri, data = df_tense_v) |> 
  distinct(verb_t, gramm, .keep_all = T) |>
  arrange(verb_t) |>  
  select(verb_t, gramm, .fitted)
```

Or we could create a list of the unique conditions.

```{r}
df_sim <-
    tibble(
    verb_t = rep(c('Past', 'Future'), each = 2),
    gramm = rep(c('0', '1'), times = 2))

# alternatively, just extract the relevant factor levels from your datafram
df_sim <-
  df_tense |> 
  arrange(verb_t) |> 
  distinct(verb_t, gramm) 

# and add predicted values
df_sim$fit <- num(predict(fit_tense_ri, df_sim), digits = 5)

df_sim
```

And now if we look at the predicted log-odds values for the future and past tenses:

```{r}
df_sim |> 
  summarise(
    mean_tense = mean(fit),
    .by = verb_t)
```

What is the difference between these two numbers (in our model summary)?

```{r}
df_sim |> 
  summarise(
    mean_tense = mean(fit),
    .by = gramm)
```

What is the difference between these two numbers (in our model summary)?

So, our slopes for `verb_t` and `gramm` correspond to the predicted difference between the two levels of each factor.

## Interpreting our coefficients

What do our coefficient estimates reflect, though? Let's remind ourselves of the rate of regressions in at the verb region:

```{r}
df_tense |> 
  filter(roi == "4") |> 
  drop_na(ri) |> 
  summary()
ptab_gramm <-
  df_tense |> 
  filter(roi == "4") |> 
  drop_na(ri) |> 
  select(gramm, ri) |> 
  table() |> 
  prop.table()

ptab_tense <-
  df_tense |> 
  filter(roi == "4") |> 
  drop_na(ri) |> 
  select(verb_t, ri) |> 
  table() |> 
  prop.table()

df_tense |> 
  filter(roi == "4") |> 
  drop_na(ri) |> 
  tabyl(gramm, ri, verb_t) |> 
  adorn_percentages() |> 
  adorn_totals()
```

We want to measure how much more likely a regression in (y = 1) is for ungrammatical conditions (x = 1) than in grammatical conditions (x = 0). Si we want to calculate the odds of a regression in for each case, and take their ratio:

```{r}
# odds(y = 1 | x = 0)
odds_ri1_gramm0 <- 
  ptab_gramm[1,2] / ptab_gramm[1,1] # in gramm conditions: ri 0/1
odds_ri1_gramm1 <- 
  ptab_gramm[2,2] / ptab_gramm[2,1] # in ungramm condiitons: ri 0/1

## odds ratio
odds_ri1_gramm1 / odds_ri1_gramm0

## log odds
log(odds_ri1_gramm1) - log(odds_ri1_gramm0)
# or
log(odds_ri1_gramm1 / odds_ri1_gramm0)

## probability
plogis(log(odds_ri1_gramm1 / odds_ri1_gramm0))
```

So the odds of a regression into the verb region is 1.4 times more likely in ungrammatical versus grammatical conditions.


```{r}
intercept <- tidy(fit_tense_ri)$estimate[1]
tense <- tidy(fit_tense_ri)$estimate[2]
gramm <- tidy(fit_tense_ri)$estimate[3]
interact <- tidy(fit_tense_ri)$estimate[4]
```

What are the log odds for the past (tense = -0.5) grammatical (`gramm` = -0.5)?

```{r}
plogis(intercept)
plogis(tense)
plogis(gramm)
plogis(interact)
```


```{r}
tidy(fit_tense_ri) |> 
  mutate(prob = plogis(estimate))
```


```{r}
plogis(intercept + tense*-.5 + gramm*-.5)
```

And past ungrammatical (gramm = +0.5)?

```{r}
plogis(intercept + tense*-.5 + gramm*.5)
```

And for the future conditions?

```{r}
plogis(intercept + tense*.5 + gramm*-.5)
```

And past ungrammatical (gramm = +0.5)?

```{r}
plogis(intercept + tense*.5 + gramm*.5)
```

```{r}
plogis(-1.22521)
```

```{r}
plogis(-1.22521)
```


\begin{align}
p &= \frac{odds}{1 + odds} \label{eq1}\\
odds &= \frac{p}{1-p} \label{eq2}\\
log\;odds &= exp(odds) \label{eq3}
\end{align}

## Visualising model predictions

Something we haven't really covered is how to visualise our model predictions. So far we've only visualised the raw data, but when interpreting model results it helps to see the predictions. This is especially true for logistic regression, because our estimates are given in log odds, which are not very intuitive.

We can use the `sjPlot` package, which is very handy:

```{r}
library(sjPlot)

plot_model(fit_tense_ri)
```

```{r}
plot_model(fit_tense_ri, type = "eff",
           terms = "verb_t")
```

```{r}
plot_model(fit_tense_ri, type = "eff",
           terms = "gramm")
```

```{r}
plot_model(fit_tense_ri, type = "int")
```

Or we can use the `ggeffects` package to extract summaries of effects, and then feed them into `ggplot2`.

```{r}
library(ggeffects)
```


```{r}
ggeffect(fit_tense_ri)

ggeffect(fit_tense_ri,
         terms = c("gramm", "verb_t"))
```





## Reporting

@sonderegger_regression_nodate-1 (Section 6.9) says the following:

> Reporting a logistic regression model in a write-up is generally similar to reporting a linear regression model...Reporting a logistic regression model in a write-up is generally similar to reporting a linear regression model: the guidelines and rationale in section 4.6 for reporting individual coefficients and the whole model hold, with some adjustments.

We can produce such a table using the `papaja` package, as in @tbl-glm-summary.

```{r}
#| label: tbl-glm-summary

library(papaja)

fit_tense_ri |> 
  apa_print() |>
  apa_table(caption = "Model summary for regressions in at the verb region. Estimates are given in log odds.")
```

Or by extracting the model summary with `tidy()`, and even adding our probabilities, as in @tbl-glm-summary2.

```{r}
#| label: tbl-glm-summary2
tidy(fit_tense_ri) |> 
  mutate(prob = plogis(estimate)) |> 
  relocate(prob, .after = std.error) |> 
  apa_table()
```


## Summary

-   we saw that the equation for a straight line boils down to its intercept and slope

-   we fit our first linear model with a categorical predictor

### Important terms {.unnumbered .smaller}

```{r}
#| echo: false
tribble(
 ~"term", ~"description/other terms",
 
) %>% kable() %>% kable_styling()
```

## Important terms {.unnumbered .smaller}


```{r}
#| echo: false
content <- 
  googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing")

content |> 
  filter(`Lecture topic` == "06 - Logistic regression") |> 
  select(-`Lecture topic`) |> 
  gt() 
```

## Learning Objectives {.unnumbered .unlisted}

Today we learned...

- how to model binomial data with logistic regression
- how to interpret log-odds and odds ratio

## Task {.unnumbered}

### Regressions out

Using the same dataset, run a logistic model exploring regressions in (`ri`) at the adverb region (`roi = "2"`). Before you run the model, do you have any predictions? Try plotting the regressions in for this region first, and generate some summary tables to get an idea of the distributions of regressions in across conditions.

### Dutch verb regularity

Load in the `regularity` data from the `languageR` package.

```{r}
df_reg <-
  regularity |> 
  clean_names()
```

> Regular and irregular Dutch verbs and selected lexical and distributional properties.

Our relevant variables will be:

- `written_frequency`: a numeric vector of logarithmically transformed frequencies in written Dutch (as available in the CELEX lexical database).
- `regularity`: a factor with levels irregular (1) and regular (0).
- `verb`: a factor with the verbs as levels.

1. Fit a logistic regression model to the data which predicts verb regularity by written frequency. Consider: What type of predictor variable do you have, and what steps should you take before fitting your model?

2. Print the model coefficients, e.g., using `tidy()`.

3. Interpret the coefficients, either in log-odds or probabilities. Report your findings.


## Session Info {.unnumbered visibility="uncounted"}

```{r}
#| eval: false
#| echo: false
RStudio.Version()$version
```


Developed with Quarto using `r R.version.string` (`r R.version$nickname`) and RStudio version 2023.9.0.463 (Desert Sunflower), and the following packages:

```{r}
sessionInfo()
```

## References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


