---
title: "Regression for Linguists"
subtitle: "WiSe23/24"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-10-10
format: 
  pdf:
    output-file: 00-welcome.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = F, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r, eval = T, cache = F}
# Create references.json file based on the citations in this script
# rbbt::bbt_update_bib("00-welcome.qmd")
```

# Welcome!

-   the language of instruction is English
-   Block course:
    -   October 10, 11, 12 (10am-4pm)
    -   January 12th (4-8pm)
    -   January 26th (4-8pm)
    -   February 9th (4-8pm)
    
# Course description

- via hands-on exercises in RStudio with the R programming language, we will develop skills and know-how related to
  + analysing linguistic data with *linear models*
  + the *theory* behind linear models
  + implement *(generalised) linear (mixed) models*
  + learn how to **communicate** our findings
    
# Course materials

- Moodle: [Regression for linguists](https://moodle.hu-berlin.de/course/view.php?id=121978)
  + Enrolment Key: reg4ling

- [Course website](https://daniela-palleschi.github.io/reg4ling) (work-in-progress)
  + https://daniela-palleschi.github.io/reg4ling

- communal terms and concepts [Google Sheet](https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing)
  + https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing
  + please keep this open during class, and write down any words/concepts that you think are important
  + you don't need to write the definition, this is a collaborative document
  + by keeping the 'Lecture topic' column accurate I can easily update the website with our terms and definitions
  
# Course credits

This course is part of [Modul 9 (Forschungsmodul)](https://gremien.hu-berlin.de/de/amb/2018/57/57_2018_ma_linguistik_2018_druck.pdf) and is worth 4 LP. A break down of the course credits:

```{r}
#| echo: false
dplyr::tribble(
  ~"LP", ~"type", ~"task",
  "1LP","preparation", "readings",
  "1LP","participation", "lecture materials/exercises",
  "1LP", "preparation", "homework due December 20th",
  "1LP", "preparation", "final homework due March 29th"
) |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```

- Office hours: Wednesdays, 15.00-16.00 (by appointment)

# Reading and preparation

- this course mainly follows @winter_statistics_2019
- by now, you should be familiar with the concepts in Chaters 1-3
  + Intro to R
  + the tidyverse and reproducible workflow
  + Descriptive statistics, models, and distributions
  
## Syllabus



## Further readings

- there are lots of useful resources out there, specifically:
  + Bodo Winter's tutorials on linear (mixed) models [@winter_linear_2013; @winter_very_2014]
  + @sonderegger_regression_nodate-1: a recently published book, I'd say more intermediate than @winter_statistics_2019
  + the [PsyTeachR](https://psyteachr.github.io/) website is a *great* resource for hands-on stats and/or data analysis in R from the University of Glasgow School of Psychology and Neuroscience
  

# A word on reproducibility {.unlisted .unnumbered visibility="uncounted"}

- I have no expectations in this course for whether you use .R scripts, Rmarkdown, or Quarto (all of which available under `File > New File`)
- If you're using Rmarkdown or Quarto (recommended), I suggest you have the following code at the end of every script:

```{r}
#| eval: false
#| echo: true
sessionInfo()
```

- this will print your session info at the end of each document to print your current packages and R version (etc.)
- In this course you'll see some examples of tools for a reproducible workflow
  + but I won't explicitly discuss them (unless it's requested)
  
# Session Info

```{r}
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
