---
title: "Equation of a line"
subtitle: "WiSe23/24"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-10-10
format: 
  pdf:
    output-file: 01-straight-line.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    execute:
      fig-align: center
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = F, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r, eval = T, cache = F}
# load packages
pacman::p_load(
  tidyverse,
  knitr,
  kableExtra,
  googlesheets4,
  gt
)

# tell googlesheets4 we don't want private
gs4_deauth()
```

## Learning Objectives {.unnumbered}

Today we will learn...

- the equation of a line
- about intercepts, slopes, and residuals

## Resources  {.unnumbered}

- relevant readings:
  + @winter_linear_2013
  + @winter_statistics_2019 (Ch. 3)

### Statistical tests versus models

- you're probably familiar with statistical tests like the *t*-test or Chi-squared test
- however, [common statistical tests are simplified linear models](https://lindeloev.github.io/tests-as-linear/) (see also [Statistical tests vs. linear regression](https://yury-zablotski.netlify.app/post/statistical-tests-vs-linear-regression/))
  + but without the added power of linear models (e.g., multiple predictors, crossed random effects)
- statistical tests tell us something about our data
- statistical *models* can generalise beyond our data

## (Linear) Regression

- we need to fit a model to our data to make predictions about hypothetical observations
  + i.e., to *predict* values of our outcome/response variable based on one (or more) predictor variables
- this model can then *predict* values of our DV based on one (or more) IV(s), i.e., *predicting* an outcome variable - because we're making predictions, we need to take into account the variability (i.e., *error*) in our data
- but how do we fit these models, and what does that even mean?

### Types of regression

```{r}
#| echo: false
tribble(
  ~"regression type", ~"predictor", ~"outcome",
  "simple regression", "Single predictor", "continuous (numerical)",
  "multiple regression", "multiple predictor", "continuous (numerical)",
  "hierarchical/linear mixed models/linear mixed effect models", "include random effect", "continuous (numerical)",
  "generalised linear (mixed) models: logistic regression", "as above","binary/binomial data",
  "generalised linear (mixed) models: poisson regression", "as above","count data"
) %>% 
  kable() %>% 
  kable_styling()
```

## Straight lines

-   *linear regression* summarises the data with a straight line
    + we *model* our data as/fit our data to a straight line
-   *straight lines* can be defined by
    -   Intercept ($b_0$)
        -   value of $Y$ when $X = 0$
    -   Slope ($b_1$)
        -   gradient (slope) of the regression line
        -   direction/strength of relationship between $x$ and $y$
        -   regression coefficient for the predictor
-   so we need to define an intercept and a slope


### A line = intercept and slope

- a line is defined by its intercept and slope
  + in a regression model, these two are called `coefficients`

```{r echo = F, fig.align = "center"}
#| label: fig-Winter-slopes-intercepts
#| fig-cap: "Image source: @winter_statistics_2019 (all rights reserved)"
#| fig-cap-location: bottom
magick::image_read(here::here("media/Winter_2019_slopes-intercepts.png"))
```

### Equation of a line

- the following are all different ways to say that a value of $y$ for a given value of $x$ (indicated by $_i$) is equal to the $intercept$ ($b_0$) plus the $slope$ ($b_1$) multiplied by the value of $x$

$$
\begin{align}
y & = mx + c\\
Y_i & = b_0 + b_1X_i \\
outcome_i & = (model) \\
y_i & = intercept + slope*x_i
\end{align}
$$

- with this equation, we can *predict* values of $y$ (our outcome variable) for a given value of $x$ (our predictor variable)

### Intercept ($b_0$)

- the value of $y$ when $x = 0$

```{r}
#| echo: false
x <- 0:9
e <- rnorm(10,mean=2,sd=1)
y <- x*.5+e
# to keep it exact:
y <- c(0.7903944 , 3.6750268 , 1.2317196 , 2.9814889 , 3.5957417 , 5.5563679 , 3.5359822 , 5.4734527 , 6.8202345 , 8.6349594)

df_random <- cbind(x,y) %>% 
  as_data_frame() %>% 
  mutate(cor_x_y = cor(x,y))

fit_random <- lm(y~x, data = df_random)
df_random$predicted <- predict(fit_random)
fit_random$residuals <- residuals(fit_random)
```

```{r}
#| echo: false
df_random %>%
  ggplot(aes(x = x, y = y)) +
  geom_abline(aes(intercept=coef(fit_random)["(Intercept)"],
                  slope=coef(fit_random)["x"]),
              colour = "blue") +
  geom_point(aes(x=0,
                  y=coef(fit_random)["(Intercept)"]),
             pch=21, 
             size=5,
             colour="red") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(limits = c(0, 9), breaks = seq(0,9,by=1)) +
  scale_y_continuous(limits = c(0, 8))
```

### Slopes ($b_1$)

::: columns
::: {.column width="70%"}
-   slopes describe a change in $y$ ($\Delta y$) over a change in $x$ ($\Delta x$)
    -   positive slope: as $x$ increases, $y$ increases
    -   negative slope: as $x$ increases, $y$ decreases
    -   if the slope is 0, there is no change in $y$ as a function of $x$
-   or: the change in $y$ when $x$ increase by 1 unit
    -   sometimes referred to as "rise over run": how do you 'rise' in $y$ for a given 'run' in $x$?
:::

::: {.column width="30%"}
$$
slope = \frac{\Delta y}{\Delta x}
$$
:::
::::

---


::: {.column width="40%"}
- what is the intercept of this line?
- what is the slope of this line?
:::

::: {.column width="60%"}
```{r}
#| echo: false
df_random %>%
  ggplot(aes(x = 3*(x-3), y = 2*(y))) +
  # geom_point(alpha = .6) +
  geom_smooth(method = "lm", se = F, alpha = .2) +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + theme(text = element_text(size = 20))           
```
:::



## Error and residuals

- *fixed effects* (IV/predictors): things we can understand/measure 
- *error* (random effects): things we cannot understand/measure
  + in biology, social sciences (and linguistic research), there will always sources of random error that we cannot account for
  + random error is less an issue in e.g., physics (e.g., measuring gravitational pull)
- *residuals*: the difference (vertical difference) between **observed data** and the **fitted values** (predicted values)

::: callout-tip

### Equation of a line

$$
\begin{align}
y & = mx + c\\
Y_i &= (b_0 + b_1X_i) + \epsilon_i\\
outcome_i & = (model) + error_i\\
y_i & = (intercept + slope*x_i) + error_i
\end{align}
$$
:::



```{r}
#| echo: false
df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "A line") +
  # geom_point(alpha = .6) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_abline(aes(intercept=coef(fit_random)["(Intercept)"],
                  slope=coef(fit_random)["x"]),
              colour = "blue") +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(limits = c(0,9.5), breaks = seq(0,9,by=1)) +
  scale_y_continuous(limits = c(0,9),expand = c(0, 0),breaks = seq(-2,9,by=1))


```

---

```{r}
#| echo: false
df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "A line with data points") +
  geom_point(alpha = .6, shape = 17, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_abline(aes(intercept=coef(fit_random)["(Intercept)"],
                  slope=coef(fit_random)["x"]),
              colour = "blue") +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(limits = c(0,9.5), breaks = seq(0,9,by=1)) +
  scale_y_continuous(limits = c(0,9),expand = c(0, 0),breaks = seq(-2,9,by=1))
```

---

```{r}
#| echo: false

df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "A line with data points and regression line") +
  geom_point(alpha = .6, shape = 17, size = 2) +
  geom_point(aes(y = predicted), colour = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_abline(aes(intercept=coef(fit_random)["(Intercept)"],
                  slope=coef(fit_random)["x"]),
              colour = "blue") +
  labs(x = "Word length",
       y = "Reading time") +
  geom_segment(aes(xend = x, yend = predicted), colour = "red") +
  theme_bw() + 
  scale_x_continuous(limits = c(0,9.5), breaks = seq(0,9,by=1)) +
  scale_y_continuous(limits = c(0,9),expand = c(0, 0),breaks = seq(-2,9,by=1))
```


### Method of least squares

- so how is any given line chosen to fit any given data?
- the ***method of least squares***
  + take a given line, and square all the residuals (i.e., $residual^2$)
  + the line with the lowest ***sum of squares*** is the line with the best fit to the given data
  + why do we square the residuals before summing them up?
    + so all values are positive (i.e., so that negative values don't cancel out positive values)
- this is how we find the ***line of best fit***
  + R fits many lines to find the one with the best fit
  
```{r}
#| echo: false

fig_point <- df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "Observed values") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  # geom_smooth(method = "lm", se = F, colour = "red") +
  # geom_segment(aes(xend = x, yend = predicted), colour = "red") +
  # geom_segment(aes(xend = x, yend = (x*.5)+1), colour = "cadetblue2") +
  # geom_abline(slope=.5, intercept=1, colour = "lightblue", size = 1) +
  geom_point(alpha = .3) +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(breaks = seq(0,10,by=1)) +
  scale_y_continuous(breaks = seq(-2,9,by=1))

fig_red <- df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "Line of best fit") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_smooth(method = "lm", se = F, colour = "red") +
  geom_segment(aes(xend = x, yend = predicted), colour = "red") +
  geom_point(alpha = .3) +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(breaks = seq(0,10,by=1)) +
  scale_y_continuous(breaks = seq(-2,9,by=1))

slope_blue <- 0.5
intercept_blue <- 1.5

fig_blue <- df_random %>%
  ggplot(aes(x = x, y = y)) +
  labs(title = "A different line") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_smooth(method = "lm", se = F, colour = "pink") +
  geom_segment(aes(xend = x, yend = (x*slope_blue)+intercept_blue), colour = "cadetblue") +
  geom_abline(slope=slope_blue, intercept=intercept_blue, colour = "cadetblue", size = 1) +
  geom_point(alpha = .3) +
  labs(x = "Word length",
       y = "Reading time") +
  theme_bw() + 
  scale_x_continuous(breaks = seq(0,10,by=1)) +
  scale_y_continuous(breaks = seq(-2,9,by=1))
```  

---

```{r}
#| label: fig-least-squares
#| fig-cap: Observed values (A), Residuals for line of best fit (B), A line of worse fit with larger residuals (C)
#| out-width: "100%"
#| fig-asp: .4
#| echo: false
library(patchwork)
fig_point + fig_red + fig_blue + plot_annotation(tag_levels = "A")
```

## Learning Objectives {.unnumbered .unlisted}

Today we learned...

## Important terms {.unnumbered .smaller}

```{r}
#| echo: false
content <- 
  googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing")

content |> 
  filter(`Lecture topic` == "01 - Equation of a line") |> 
  select(-`Lecture topic`) |> 
  gt() 
```

## Exercise

### Pen-and-paper

You will receive a piece of paper with several grids on it. Follow the instructions, which include drawing some lines.

## Literaturverzeichnis {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
