---
title: "Multiple regression"
subtitle: "WiSe23/24"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-10-11
format: 
  pdf:
    output-file: 04-multiple_regression_slides.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    output-location: fragment
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```


# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

- what multiple regression is
- how to include multiple predictor variables
- how to interpret slopes in multiple regression
- how to interpret interaction effects
- about the assumption of the absence of collinearity

# Set-up environment  {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
```

```{r}
# load packages
pacman::p_load(
               tidyverse,
               here,
               broom,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra,
               gt,
               googlesheets4)

# tell googlesheets4 we don't want private
gs4_deauth()
```

## Load data {.unnumbered .unlisted}

- we'll use the full dataset of the frequency data.

```{r}
df_freq_full <-
  read_csv(here("data", "ELP_full_length_frequency.csv")) |> 
  clean_names() |> 
  mutate(freq = 10^(log10freq), # inverse log10
         freq_log = log(freq)) |>  # use natural logarithm
  relocate(word, rt, length, freq, freq_log)
  
```

- the variable `log10freq` is a remnant from @winter_statistics_2019; we'll use the natural logarithm

# Multiple regression

- so far we've run simple linear models, which are equivalent to
  + a one-sample *t*-test (intercept-only model)
  + two-sample *t*-test (for a categorical predictor)
  + Pearson's *r* (for a standardised continuous predictor)
- why then should we bother with linear regression?
  + it allows us to include *multiple* predictors in our models simultaneously
    + still boils down to modeling the mean, but while conditioning the mean on multiple variables at once

## Extending the equation of a line

- the equation of a line (@eq-simple-lin-2):
  + value of $y$ = the intercept ($b_0$) + the corresponding value of $x$ multiplied by the slope ($b_1x$) + the error (residuals ($e$))
- In multiple regression, we can include more than one slope (@eq-multiple-reg)

$$
y = b_0 + b_1x + e
$$ {#eq-simple-lin-2}

$$
y = b_0 + b_1x + b_2x + ... + e
$$ {#eq-multiple-reg}

## One predictor

- re-run our simple model with this dataset
  + keep reaction times un-transformed for now

```{r}
fit_freq_full <-
  lm(rt ~ log(freq), data = df_freq_full)
```

```{r}
tidy(fit_freq_full)
```

- there is a decrease in reaction times (-37.5 milliseconds) for a 1-unit increase in log frequency

---

- model fit using `glance()`:

```{r}
glance(fit_freq_full)$r.squared
```

- our model describes 38% of the variance in response times
- is this described variance due solely to frequency?
  + Other effects that are correlated with frequency might be conflating the frequency effect
  + Let's expand our model to include word length [@eq-freq-length]

$$
y = b_0 + b_1*log\;frequency + b_2*word\;length
$$ {#eq-freq-length}

## Adding another predictor

- add `length` as a predictor

```{r}
fit_freq_mult <-
  lm(rt ~ log(freq) + length, data = df_freq_full)
```

```{r}
tidy(fit_freq_mult) |> select(term, estimate)
```

- an increase in word length (+1 letter) corresponds to a 20ms increase in reaction times
  + our intercept is now 748ms, instead of 907ms
  + this corresponds to the prediction for reaction times to a word with 0 log frequency and 0 word length, but this is not very interpretable
  
---
  
```{r}
b0 <- tidy(fit_freq_mult)$estimate[1]
b1_freq <- tidy(fit_freq_mult)$estimate[2]
b1_length <- tidy(fit_freq_mult)$estimate[3]
```

```{r}
b0 + b1_freq*0 + b1_length*0
```

- centering both predictors would result in an intercept reflecting the reaction time for a word with average frequency and average length (because 0 for each would equal the mean)

- the slope for log frequency has also changed: from -37.5 to -29.5
  + this tells us that some of the effect in our first model was confounded with length, as controlling for length weakens the effect of frequency

```{r}
glance(fit_freq_mult)$r.squared
```

- including `length` increases the variance described by our model, reflected in the $R^2$ values (`r glance(fit_freq_mult)$r.squared` instead of `r glance(fit_freq_full)$r.squared`

# Standardising our predictors

- when we have multiple continuous predictors, standardising them can help their interpretation
  + because their slopes are comparable
- we can do this by centering each variable and then dividing by the standard deviation (like yesterday), or we could use the `scale()` function, which has the same result

```{r}
#| output-location: column-fragment
# centre and then standardize
df_freq_full |> 
  mutate(
         freq_z1 = (freq-mean(freq))/sd(freq),
         freq_z2 = scale(freq)) |> 
  select(freq_z1, freq_z2) |> 
  head()
```

---

- use `scale()` for `freq` and `length`.

```{r}
df_freq_full <-
  df_freq_full |> 
  mutate(freq_z = scale(freq_log),
         length_z = scale(length))
```

- fit a model with them as predictors

```{r}
fit_freq_z <-
  lm(rt ~ freq_z + length_z, data = df_freq_full)
```

---

First, let's check the *$R^2$*:

```{r}
#| output-location: column-fragment
glance(fit_freq_z)$r.squared
```

- *$R^2$* = `r glance(fit_freq_z)$r.squared`, just like above
  + this is a reminder that the predictors still represent the same variance in the underlying model
  + their units and scales have simply changed. 

---

- what about our coefficients:

```{r}
tidy(fit_freq_z) |> select(term, estimate)
```

- a 1-unit change now corresponds to a change of 1 standard deviation (because we standardized, producing z-scores)
- frequency now has a larger magnitude than the effect of length
  + a 1-SD increase in frequency (holiding length constant) = decrease in reaction times by 60.6 ms
  + a 1-SD increase in length (holiding frequency constant) = increase in reaction times by 43.3 ms
  + does frequency influence the effect of length, and vice versa?

## Adding an interaction term

- please check out Ch. 8 (Interations and nonlinear effects) in @winter_statistics_2019 for a more in-depth discussion on interactions
- For now, what's important to know is that interactions describe how effects of one predictor may be influenced by changes in another predictor

---

- We can add interaction terms of two predictors by connecting them with a colon (`:`).

```{r}
#| output-location: column-fragment
lm(rt ~ freq_z + length_z + freq_z:length_z, 
   data = df_freq_full) |> 
  tidy() |> select(term, estimate)
```

- Or connect the two predictors with an asterisk (`*`) to indicate that we want to look at both predictors and their interaction

```{r}
#| output-location: column-fragment
lm(rt ~ freq_z*length_z, 
   data = df_freq_full) |> 
  tidy() |> select(term, estimate)
```

- the intercept is the predicted reaction time for a word with the mean length and mean frequency
  + Notice that the interaction slope is negative, meaning when both `freq` and `length` increase, reaction times will decrease

# Model assumptions

- we've discussed the assumptions of normality and homoscedasticity (constant variance), which both refer to the residuals of a model    
  + We typically assess these assumptions visually

## Normality and Homoscedasticity

- for our model:

```{r}
#| code-fold: true
#| output-location: fragment
#| out-width: 100%
#| fig-asp: .35
fig_hist <-
fit_freq_z |> 
  ggplot() +
  aes(x = .resid) +
  geom_histogram(bins = 20, fill = "grey", colour = "black") +
  theme_bw() +
  labs(title='Histogram', x='Residuals', y='Count')

fig_qq <-
fit_freq_z |> 
  ggplot() +
  aes(sample = .resid) +
  geom_qq(colour = "red") +
  geom_qq_line() +
  labs(title='Q-Q Plot', x='Theoretical quantiles', y='Sample quantiles')

fig_res <-
  fit_freq_z |> 
  ggplot() +
  aes(x = .fitted, y = .resid) +
  geom_point() +
  geom_hline(yintercept = 0, colour = "blue") +
  labs(title='Residual vs. Fitted Values Plot', x='Fitted Values', y='Residuals')

fig_hist + fig_qq + fig_res
```

- not very reassuring, let's try log-transformed reaction times.

## Log-transformed response variable

```{r}
fit_freq_log_z <-
  lm(log(rt) ~ freq_z*length_z,
     data = df_freq_full)
```

```{r}
glance(fit_freq_log_z)$r.squared
```

```{r}
tidy(fit_freq_log_z) |> select(term, estimate)
```

- our coefficients are much smaller, because they're on the log-scale

---

```{r}
#| code-fold: true
#| output-location: fragment
#| out-width: 100%
#| fig-asp: .35
fig_hist <-
fit_freq_log_z |> 
  ggplot() +
  aes(x = .resid) +
  geom_histogram(bins = 20, fill = "grey", colour = "black") +
  theme_bw() +
  labs(title='Histogram', x='Residuals', y='Count')

fig_qq <-
fit_freq_log_z |> 
  ggplot() +
  aes(sample = .resid) +
  geom_qq(colour = "red") +
  geom_qq_line() +
  labs(title='Q-Q Plot', x='Theoretical quantiles', y='Sample quantiles')

fig_res <-
  fit_freq_log_z |> 
  ggplot() +
  aes(x = .fitted, y = .resid) +
  geom_point() +
  geom_hline(yintercept = 0, colour = "blue") +
  labs(title='Residual vs. Fitted Values Plot', x='Fitted Values', y='Residuals')

fig_hist + fig_qq + fig_res
```

- looks better

## Collinearity

- collinearity refers to when continuous predictor variables are correlated
  + can make the interpretation of their coefficients difficult, and the results spurious
- regression assumes there is an *absence* of collinearity
  + i.e., our predictor variables are not correlated.

---

- the `vif()` function from the `car` package assesses collinearity, comparing *variance inflation factors*
- VIF values close to 1 indicates there is not a high degree of collinearity between your variables
  + higher than 1 indicates correlation; above 10 is highly correlated (thresholds may be field-specific)

```{r}
#| output-location: fragment
car::vif(fit_freq_log_z)
```

---

- collinearity is a conceptual problem, and should be considered in the planning stage
- we want to include predictors that we have specific predictions or research questions about
  + shoving a bunch of predictors in a model to see what comes out significant is bad practice
- we should have a principled approach to model building and variable selection
- of course, we can add predictors in exploratory analyses, but this should always be reported as exploratory
  + and any observed effects replicated where possible

## Adjusted $R^2$

- adjusted $R^2$ is a more conservative version of $R^2$ that takes into account the number of predictors in a model
- adjusted $R^2$ includes the number of predictors ($k$) in its denominator (bottom half of a division)
  + the more predictors there are, the smaller adjusted $R^2$ will be, unless each additional predictor explains sufficient variance to counteract this penalisation

---

```{r}
#| output-location: column-fragment
glance(fit_freq_z)$adj.r.squared
glance(fit_freq_log_z)$adj.r.squared
```

- there is a small increase in adjusted $R^2$ when we include length and its interaction with frequency
  - this suggests that including `length` and the interaction term does not result in overfitting
  + i.e., length contributes to the variance explained by the model

---

- if we compare to the same model without an interaction term (log reaction times ~ frequency + length), we see that the adjusted $R^2$ is not very different. 
- if the adjusted $R^2$ were much lower, this would indicate that including the interaction term leads to overfitting

```{r}
glance(lm(log(rt) ~ freq_z + length_z, data = df_freq_full))$adj.r.squared
```


## Important terms {.unnumbered .smaller}


```{r}
#| echo: false
content <- 
  googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing")

content |> 
  filter(`Lecture topic` == "04 - Multiple regression") |> 
  select(-`Lecture topic`) |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```


# Learning Objectives {.unnumbered .unlisted}

Today we learned...

- what multiple regression is
- how to include multiple predictor variables
- how to interpret slopes in multiple regression
- how to interpret interaction effects
- about the assumption of the absence of collinearity

# Task

Load in the `english` dataset from the `languageR` package [@languageR-package] (code below). You don't need to load in any CSV file, because this dataset is available if you have the package loaded. From the manual:

> This data set gives mean visual lexical decision latencies and word naming latencies to 2284 monomorphemic English nouns and verbs, averaged for old and young subjects, with various predictor variables.

([languageR manual, p. 29](https://cran.r-project.org/web/packages/languageR/languageR.pdf))

```{r}
# load in 'english' dataset from languageR
df_freq_eng <-
  as.data.frame(english) |> 
  dplyr::select(RTlexdec, RTnaming, Word, LengthInLetters, AgeSubject, WrittenFrequency) |> 
  rename(rt_lexdec = RTlexdec,
         rt_naming = RTnaming,
         freq_written = WrittenFrequency) |> 
  clean_names() |> 
  relocate(word)
```

We're keeping five variables:

- `word`: a factor with 2284 words
- `rt_lexdec`: numeric vector of log RT in visual lexical decision
- `rt_naming`: numeric vector of log RT in word naming
- `length_in_letters`: numeric vector with length of the word in letters
- `AgeSubject`: a factor with as levels the age group of the subject: young versus old.
- `freq_written`: numeric vector with log frequency in the CELEX lexical database

Take the following steps:

1. Perform an exploratory data analysis to understand the data (produce plots, tables, whatever you think necessary and can do).

2. Model the data, with *back-transformed* (raw) reaction times as a response variable and written frequency and length in letters as predictors. Perform any tranformations you think necessary. Run model diagnostic checks and assess model fit.

```{r}
#| echo: false
#| eval: false
fit_freq_eng <-
  lm(exp(rt_lexdec) ~ freq_written * length_in_letters, data = df_freq_eng)

summary(fit_freq_eng)

performance::check_model(fit_freq_eng)
```

3. Re-run the model with log reaction times as a response variable and written frequency and length in letters as predictors. Perform any tranformations you think necessary. Run model diagnostic checks and assess model fit.


```{r}
#| echo: false
#| eval: false
fit_freq_eng_log <-
  lm(rt_lexdec ~ freq_written * length_in_letters, data = df_freq_eng)

summary(fit_freq_eng_log)
glance(fit_freq_eng)
glance(fit_freq_eng_log)
car::vif(fit_freq_eng_log)
car::vif(fit_freq_eng)

# interaction term highly corr'd with frequency

performance::check_model(fit_freq_eng_log)
```

4. Remove length in letters as a predictor. How is model fit affected? What can you conclude?

```{r}
#| echo: false
#| eval: false
fit_freq_eng_simp <-
  lm(rt_lexdec ~ freq_written, data = df_freq_eng)

summary(fit_freq_eng_simp)
glance(fit_freq_eng_simp)
glance(fit_freq_eng_log)

performance::check_model(fit_freq_eng_simp)

# decision: should length and interaction effect be included? doesn't hurt or help model fit it seems
```

# Literaturverzeichnis {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


