---
title: "Categorical variables"
subtitle: "WiSe23/24"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-10-11
format: 
  pdf:
    output-file: 04-multiple_regression_slides.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    output-location: fragment
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```


# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

- about cateogorical predictors
- how to interpret different contrast coding

# Set-up environment  {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
```

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               lme4,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra)
```

## Load data {.unnumbered .unlisted}

- load in the the dataset from the `languageR` package


```{r}
df_freq_eng <-
  as.data.frame(english) |> 
  # keep relevant variables
  dplyr::select(RTlexdec, RTnaming, Word, LengthInLetters, AgeSubject, WrittenFrequency) |> 
  # rename some variables
  rename(rt_lexdec = RTlexdec,
         rt_naming = RTnaming,
         freq_written = WrittenFrequency) |> 
  clean_names() |> 
  # standardize continuous predictors
  mutate(
    freq_z = scale(freq_written),
    length_z = scale(length_in_letters)
  ) |> 
  # move 'word' to front
  relocate(word) |> 
  # arrange alphabetically by 'word'
  arrange(word)
```

# Bimodal distribution

- in your exploratory data analysis, you might've noticed a *bimodal* distribution.

```{r}
#| echo: false
fig_hist_log <- 
  df_freq_eng |> 
  ggplot() +
  aes(x = rt_lexdec) +
  geom_histogram() 

fig_dens_log <- 
  df_freq_eng |> 
  ggplot() +
  aes(x = rt_lexdec) +
  geom_density() 


fig_hist <-
  df_freq_eng |> 
  ggplot() +
  aes(x = rt_lexdec, fill = age_subject) +
  geom_histogram(alpha = .5, position = "stack") +
  # facet_grid(.~age_subject) +
  theme(legend.position = "none")

fig_dens <-
  df_freq_eng |> 
  ggplot() +
  aes(x = rt_lexdec, colour = age_subject, fill = age_subject) +
  stat_density(alpha = .4) +
  theme(legend.position = "bottom")



  (fig_hist_log + fig_dens_log) /
  (fig_hist + fig_dens)
```

---

- this is a *bimodal* distribution
  + there are two *modes* (most frequent value, i.e., peak in a histogram)
- We know that there were two subject groups: old and young
  + it might be that each group has a different mode

---

Running our model of the log reaction times as predicted by frequency and length, we see:

```{r}
fit_freq_length <-
  lm(rt_lexdec ~ freq_z*length_z,
     data = df_freq_eng)
```

```{r}
glance(fit_freq_length)$r.squared
glance(fit_freq_length)$adj.r.squared
```

Seems like we don't have any overfitting in our model ($R^2$ and adjusted $R^2$ are comparable). Let's look at our coeffiecients.

```{r}
tidy(fit_freq_length) |> select(term, estimate)
```

- looks to the dataset we explored in the previously
- the bimodal distribution suggests we should include age group as a predictor, since the two groups seem to pattern differently in their reading times
- does the effect of frequency and length also differ as a function of age group?

# Categorical predictors

- we'd predict longer reading times for older participants than younger participants
  + although we should hypothesise *before* collecting and visualising our data!
- though age is numerical, all we have is two categories: old or young

## Including a categorical predictor

- include `age_subject` in our model

```{r}
fit_age <-
  lm(rt_lexdec ~ freq_z*length_z + age_subject,
     data = df_freq_eng)
```

```{r}
glance(fit_age)$r.squared
glance(fit_age)$adj.r.squared
```

- large increase in variance explained
- and that the $R^2$ and adjusted $R^2$ values are comparable

```{r}
car::vif(fit_age)
```

- VIF values for all coefficients are near 1
  + this indicates that our predictors all contribute to the variance explained by the model and are not correlated
  
# Contrasts

- let's take a look at our model estimates

```{r}
tidy(fit_age) |> select(term,estimate)
```

- there is a negative slope for `age_subjectyoung`
  + reaction times decrease when...what?
- how does a categorical variable get fit to a line? 
- the factor levels (i.e., the categories in a categorical variable) are given numerical values
  + We call these numerical values mapped onto factor levels contrast coding

## Dummy contrasts

- we can check the contrasts with `contrasts()`

```{r}
contrasts(df_freq_eng$age_subject)
```

- `old` was coded at $0$ and `young` as $1$
- our slope for `age_subjectyoung` represents the change in reaction times when we move from `old` to `young`
- this is called `treatment coding` (dummy coding), where one factor level is coded as 0 and the other as 1

---

- remove frequency and length to focus on `age_subject`
- use raw reaction times, to more easily interpret the results

```{r}
fit_age <-
  lm(exp(rt_lexdec) ~ age_subject,
     data = df_freq_eng)
```

```{r}
glance(fit_age)$r.squared
```

- $R^2$ is lower than when we included frequency and length
  + but higher than our model with frequeny and length but no age

---

```{r}
tidy(fit_age) |> select(term, estimate)
```

- reaction times decrease by 157ms going from old t young group compared to the old group
- what does the intercept represent here?

```{r}
#| output-location: column-fragment
df_freq_eng |> 
  select(rt_lexdec, age_subject) |> 
  mutate(rt_lexdec = exp(rt_lexdec)) |> 
  summary()
```

- don't see the intercept value there

---

- how does `rt_lexdec` look for the two groups?

```{r}
#| output-location: column-fragment
df_freq_eng |> 
  select(rt_lexdec, age_subject) |> 
  mutate(rt_lexdec = exp(rt_lexdec)) |> 
  summarise(mean = mean(rt_lexdec),
            min = min(rt_lexdec),
            max = max(rt_lexdec),
    .by = "age_subject"
  )
```

- the intercept corresponds to the mean reaction time for the old group. Why? 
  + because `old` coded as $0$

---

- which variable is coded as $0$? 
  + R simply takes the first level name alphabetically: `old` comes before `young`, so `old` was automatically taken as the 'baseline' to which `young` was compared


- if we were to add the slope to the intercept, we would get the mean for the $young$ group. Why is this?

```{r}
coef(fit_age)['(Intercept)'] + coef(fit_age)['age_subjectyoung']
```

---

- this actually is the same thing as a *t*-test:

```{r}
t.test(exp(rt_lexdec) ~ age_subject, data = df_freq_eng)
```

- if we compare this to our model, we see that the *t*- and *p*-values are identical (more on these later).

```{r}
tidy(fit_age)
```

---

```{r}


fig_nocontrasts <-
df_freq_eng |> 
  ggplot() +
  aes(x = age_subject, y = exp(rt_lexdec)) +
  labs(title = "No contrasts") +
  # geom_vline(xintercept = 0, linetype="dashed", size = .5) +  
  geom_point(position = position_dodge(.6)) + 
  geom_smooth(method = 'lm', aes(group=1)) + theme_minimal() +
  theme_bw()

fig_treatment <-
df_freq_eng |> 
  mutate(age_subject = if_else(age_subject=="young",1,0)) |>
  ggplot() +
  aes(x = age_subject, y = exp(rt_lexdec)) +
  labs(title = "Treatment contrasts") +
  geom_vline(xintercept = 0, linetype="dashed", size = .5) +
  geom_point(position = position_dodge(.6)) + 
  geom_smooth(method = 'lm', aes(group=1)) + theme_minimal() +
  theme_bw()

fig_nocontrasts + fig_treatment
```


## Sum contrasts

- sum coding is another frequently used coding scheme
  + essentially centring categorical variables
- simplifies interpretation of interaction effects
- instead of $0$ and $1$, we set our contrasts to +/-1 or 0.5 (I prefer 0.5)

---

```{r}
# first, make sure your variable is a factor
df_freq_eng$age_subject <- as.factor(df_freq_eng$age_subject)
# check
class(df_freq_eng$age_subject)
```

---

```{r}
# next, you could use the contr.sum() function
contrasts(df_freq_eng$age_subject) <- contr.sum(2) # where 2 means we have 2 levels
contrasts(df_freq_eng$age_subject)
```

- `old` is coded as $-1$ and `young` as $+1$
- I prefer to use +/-0.5 for reasons we don't need to go into here
  + I would also prefer to have `young` coded in the negative value, and `old` in the positive value
  + this aids in the way I interpret the slope: a change in reaction times for the older group compared to the younger group
  
---

```{r}
#or, you could manually control the sum contrasts
# check the order of the levels
levels(df_freq_eng$age_subject)

# code 'old' as +.5 and 'young' as -.5
contrasts(df_freq_eng$age_subject) <- c(+0.5, -0.5)
contrasts(df_freq_eng$age_subject)
```

- run our model

```{r}
fit_age_sum <-
  lm(exp(rt_lexdec) ~ age_subject,
     data = df_freq_eng)
```

```{r}
glance(fit_age_sum)$r.squared
glance(fit_age)$r.squared
```

- no difference in variance account for by our model (remember, centering a variable just shifts values, doesn't affect the relationship between values)

---

```{r}
tidy(fit_age_sum) |> select(term,estimate)
```

- there is a difference in the intercept
  + andd a change in sign in our slope. Why is this?

```{r}
#| label: fig-contrasts
#| output-location: fragment
#| code-fold: true
#| fig-cap: The difference in slope corresponds to which level is coded as 0 (dummy coding) or -5/-1 (sum coding)
fig_sum1 <-
df_freq_eng |> 
  mutate(age_subject = if_else(age_subject=="young",-1,1)) |>
  ggplot() +
  aes(x = age_subject, y = exp(rt_lexdec)) +
  labs(title = "Sum contrasts") +
  geom_vline(xintercept = 0, linetype="dashed", size = .5) +
  geom_point(position = position_dodge(.6)) + 
  geom_smooth(method = 'lm', aes(group=1)) + theme_minimal() +
  theme_bw()

fig_sum5 <-
df_freq_eng |> 
  mutate(age_subject = if_else(age_subject=="young",-.5,.5)) |>
  ggplot() +
  aes(x = age_subject, y = exp(rt_lexdec)) +
  labs(title = "Sum contrasts") +
  geom_vline(xintercept = 0, linetype="dashed", size = .5) +
  geom_point(position = position_dodge(.6)) + 
  geom_smooth(method = 'lm', aes(group=1)) + theme_minimal() +
  theme_bw()

fig_treatment + fig_sum5 + plot_annotation(tag_levels = "A")
```

---

- the intercept value is now the overall mean of all observed reaction times, because now the $y$ value when $x$ equals zero lies in the middle of the two groups
- the slope magnitude (i.e., size of the value) hasn't changed, because the difference betwen the two group means has not changed

```{r}
mean(exp(df_freq_eng$rt_lexdec))
```


# Exploring predicted values

- let's explore the predicted values of our model with a categorical variable

```{r}
head(fitted(fit_age), n = 10)
```

- there are only 2 values, 630 and 787
  + these correspond to the means for each group that we saw above
  + they also seem to be in a pattern: mean(young), mean(old), mean(young), mean(old), etc.
  + how does this correspond to the age group of the participant for the first ten observations?

```{r}
head(df_freq_eng$age_subject, n = 10)
```

---

- first ten observations in our data are in young-old pairs. What are the first values in the raw data?

```{r}
head(exp(df_freq_eng$rt_lexdec), n = 10)
```

- what is the difference between these reaction times and the fitted values?

```{r}
head(exp(df_freq_eng$rt_lexdec), n = 10) - head(fitted(fit_age), n = 10)
```

```{r}
head(residuals(fit_age))
```

---

- we see again that predicted values correspond to the $x$ value for the corresponding row in the dataframe
  + but with our two-level factor, we only have two $x$ values, young and old
  
---

```{r}
df_freq_eng <- 
  augment(fit_age, df_freq_eng)
```

```{r}
df_freq_eng |> 
  select(word, age_subject, rt_lexdec, .fitted, .resid) |> 
  mutate(rt_lexdec = exp(rt_lexdec)) |> 
  head()
```




  

# Summary

-   we saw that the equation for a straight line boils down to its intercept and slope

-   we fit our first linear model with a categorical predictor

## Important terms {.unnumbered .smaller}

```{r}
#| echo: false
tribble(
 ~"term", ~"description/other terms",
 
) %>% kable() %>% kable_styling()
```


# Learning Objectives {.unnumbered .unlisted}

Today we learned...


# Task



# Literaturverzeichnis {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


