---
title: "Mixed Models 1"
subtitle: "Random intercepts"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-01-12
format: 
  pdf:
    output-file: 08-lmm1_random_intercepts.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    include-in-header: ../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    output-location: fragment
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn...



# Resources {.unnumbered .unlisted}

- this lecture covers
  + Sections 
  
# Set-up {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
options(pillar.sigfig = 5)
```

```{r}
#| echo: false
library(broman)
# function to format p-values
format_pval <- function(pval){
	dplyr::case_when(
		pval < .001 ~ "< .001",
		pval < .01 ~ "< .01",
		pval < .05 ~ "< .05",
		TRUE ~ broman::myround(pval, 3)
	)
}
```

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra)
```

```{r}
# set preferred ggplot2 theme
theme_set(theme_bw() + theme(plot.title = element_text(size = 10)))
```

### Load packages {.unnumbered}

We'll also need to load in our required packages. Hopefully you've already install the required packages (if not, go to @sec-software).

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               janitor,
               ggeffects,
               sjPlot,
               # new packages:
               lme4,
               lmerTest,
               broom.mixed)
```

Here I also globally set my preferred `ggplot2` theme so that all of my plots are formatted how I like them, without have to repeat the code for each plot. This is completely optional.

```{r}
# set preferred ggplot2 theme
theme_set(theme_bw() + theme(plot.title = element_text(size = 10)))
```

#### Resolve conflicts  {.unnumbered}

- both `lme4` and `lmerTest` have a function `lmer()`   
  + for now we want to use the `lme4` version

```{r}
lmer <- lme4::lmer
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra,
               googlesheets4,
               gt)

# tell googlesheets4 we don't want private
gs4_deauth()
```


### Load data {.unnumbered}

- data from @biondo_yesterday_2022

```{r}
df_biondo <-
  read_csv(here("data", "Biondo.Soilemezidi.Mancini_dataset_ET.csv"),
           locale = locale(encoding = "Latin1") ## for special characters in Spanish
           ) |> 
  clean_names() |> 
  mutate(gramm = ifelse(gramm == "0", "ungramm", "gramm")) |> 
  mutate_if(is.character,as_factor) |> # all character variables as factors
  filter(adv_type == "Deic")
```

And take a look at the data:

```{r}
head(df_biondo)
```

### Sum contrasts

```{r}
#| output-location: column-fragment
contrasts(df_biondo$verb_t)
contrasts(df_biondo$gramm)
```

```{r}
contrasts(df_biondo$verb_t) <- c(-0.5,+0.5)
contrasts(df_biondo$gramm) <- c(-0.5,+0.5)
```


```{r}
#| output-location: column-fragment
contrasts(df_biondo$verb_t)
contrasts(df_biondo$gramm)

```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

# Genearlised linear mixed models

- e.g., logistic regression (`family = "binomial"`)
- the same logic as linear (mixed) regression
- let's take a look at the regressions out in the verb region
  - N.B., there's no `glmer()` function in `lmerTest`, we have to use `lme4`

## 

```{r}
#| warning: true
#| message: true
fit_glmm_ro_sj_item <-
  lme4::glmer(ro ~ verb_t*gramm +
                    (1|sj) +
                    (1|item),
                  data = df_biondo,
                  subset = roi == 4,
        family = "binomial"
        )
```

- we get a warning that our model failed to converge
- let's check out summary output

##

```{r}
summary(fit_glmm_ro_sj_item)
```

- the model output looks very similar to that for `lmer()`
- our effects look really big (high t-values, small p-values)
  + this seems unlikely
- we also get a warning message at the very bottom: Model failed to converge
  + we cannot report a model that fails to converge

### Set optimizer and increase interations

```{r}
fit_glmm_ro_sj_item <-
  lme4::glmer(ro ~ verb_t*gramm +
                    (1|sj) +
                    (1|item),
                  data = df_biondo,
                  subset = roi == 4,
        family = "binomial",
        # this line is used because we get a convergence warning without it:
        glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
        # we increase the number of iterations and set the optimizer to bobyqa
        )
```

- no warnings, so we can report this model

##

```{r}
summary(fit_glmm_ro_sj_item)
```

- we see firstly that we get the same information, plus measures of model fit (AIC, BIC, logLik)
- our model estimates are in log odds
  + `> 0` = regression in more likely than not
  + `< 0` = regression in less likely than not
- note also that we have *z*-values instead of *t*-values (this is always the case with logistic regression)

## Convergence warnings

- convergence warnings are very common when fitting linear mixed models in the Frequentist (i.e., not Bayesian) framework
- some common strategies:
  + try alternative optimizers
  + increase iteration count
  + failing this, you may need to simplify your random effects structure
- convergence issues are the reason why many reported findings come from random intercepts-only models
  + these models can still have an inflated Type I error rate
  + the solution to this? Add random effects...which sometimes cause convergence issues...
- the real solution: go Bayesian, or try Julia...

# Learning objectives {.unnumbered .unlisted .uncounted}

# Important terms {.unnumbered .smaller .uncounted}

```{r}
#| echo: false
tribble(
 ~"term", ~"description/other terms",
 
) %>% kable() %>% kable_styling()
```

# Task

1. Fit a linear model (`lm()` function) to log-transformed total reading times (`tt`) at the adverb region (`roi == 2`), with adverb time reference (`adv_t`) and gramm (`gramm`) and their interaction as fixed effects. Use sum contrast coding (`Past` and `gramm` = `-0.5`, `Future` and `ungramm` = `+0.5`). Save this model as `fit_lm_adv_tt`.

2. Fit a linear mixed model similar to `fit_lm_adv_tt` by including by-participant and by-item varying intercepts. Save this model as `fit_lmm_adv_tt`.

3. Inspect the fixed effects for both models. How do they differ? Describe the observed coefficients.

5. Plot the fixed effects for `fit_lm_adv_tt` and `fit_lmm_adv_tt`.

4. Inspect the random effects for `fit_lmm_adv_tt`. What do you see?

6. Plot the random effects per participant and item.

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


