---
title: "Mixed Models 1"
subtitle: "Random intercepts"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-01-12
format: 
  pdf:
    output-file: 08-lmm1_random_intercepts.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    include-in-header: ../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

- what linear mixed models are
- how to fit a random-intercepts model
- how to inspect and interpret a mixed effects model


# Resources {.unnumbered .unlisted}

- this lecture covers
  + Chapter 14 \'Mixed Models 1: Conceptual Introduction' [until Section 14.8\; @winter_statistics_2019]
  + @winter_very_2014 (until page 16)
  + Sections 8.1-8.3 in @sonderegger_regression_nodate-1
- we will be using the data from @biondo_yesterday_2022
  
# Set-up {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
```

```{r}
#| code-fold: true
#| code-summary: Code for a function to format p-values
library(broman)
# function to format p-values
format_pval <- function(pval){
	dplyr::case_when(
		pval < .001 ~ "< .001",
		pval < .01 ~ "< .01",
		pval < .05 ~ "< .05",
		TRUE ~ broman::myround(pval, 3)
	)
}
```

## Load packages {.unnumbered}

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               janitor,
               ggeffects,
               sjPlot,
               # new packages for mixed models:
               lme4,
               lmerTest,
               broom.mixed,
               lattice)
```

```{r}
#| echo: false
# set preferred ggplot2 theme
theme_set(theme_bw() + theme(plot.title = element_text(size = 10)))
```

### Resolve conflicts  {.unnumbered}

- both `lme4` and `lmerTest` have a function `lmer()`   
  + for now we want to use the `lme4` version

```{r}
lmer <- lme4::lmer
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra,
               googlesheets4,
               gt)

# tell googlesheets4 we don't want private
gs4_deauth()
```


## Load data {.unnumbered}

- data from @biondo_yesterday_2022

```{r}
df_biondo <-
  read_csv(here("data", "Biondo.Soilemezidi.Mancini_dataset_ET.csv"),
           locale = locale(encoding = "Latin1") ## for special characters in Spanish
           ) |> 
  clean_names() |> 
  mutate(gramm = ifelse(gramm == "0", "ungramm", "gramm")) |> 
  mutate_if(is.character,as_factor) |> # all character variables as factors
  filter(adv_type == "Deic") |> 
  droplevels()
```

And take a look at the data:

```{r}
head(df_biondo)
```

## Set contrasts


```{r}
contrasts(df_biondo$verb_t) <- c(-0.5,+0.5)
contrasts(df_biondo$gramm) <- c(-0.5,+0.5)
```


```{r}
#| output-location: column-fragment
contrasts(df_biondo$verb_t)
```

```{r}
#| output-location: column-fragment
contrasts(df_biondo$gramm)
```

# Linear mixed (effects) models

- mixed models allow for varying intercepts and slopes per level of some grouping factor
- recall that intercepts (can) represent the grand mean of the data
- slopes represent a change in $y$ for a 1-unit change in $x$ ($\frac{\Delta y}{\Delta x}$, "rise over run")
  + i.e., the difference between two categories, or for a 1-unit change of a continuous predictor

- random intercepts take into account that each level of a grouping factor can vary in their mean
- random slopes take into account that each level of a grouping factor can vary in the effect of a predictor

## Random intercepts vs. random slopes

- @biondo_yesterday_2022 used a within-participant, a.k.a. repeated measures design
  + `r length(unique(df_biondo$sj))` participants saw `r length(unique(df_biondo$item))` items, rotated throughout the conditions in a Latin square design

- we would expect some participants to be faster readers than others
  + this would be reflected in a shorter mean reading time
- some participants will tend to have a stronger effect of e.g., grammaticality than others
  + this would be reflected in a steeper slope for `gramm`aticality
- the same could be said for certain experimental items
  + reading times will vary by item e.g., for word length or familiarity
  + some items will also tend to have a stronger effect than others

### By-participant variance

```{r}
#| echo: false
#| label: fig-boxplot-sj
#| fig-cap: By-participant boxplot of first-pass RTs at the verb region with overall median FP value in red
df_biondo |> 
  filter(roi == "4") |> 
  mutate(sj_median = median(fp, na.rm = T), .by=sj) |> 
  arrange(desc(sj_median)) |> 
  ggplot() +
  aes(x = reorder(sj, sj_median),
      y = fp) +
  labs(title = "By-participant first-pass RTs at the verb",
       y = "First-pass reading times (ms)",
       x = "Participant ID") +
  geom_hline(
    colour = "red", size = 1,
    aes(yintercept = median(fp, na.rm = T))
  ) +
  geom_boxplot(alpha = 0.5) 
```

### By-item variance

```{r}
#| echo: false
#| label: fig-boxplot-item
#| fig-cap: By-item boxplot of first-pass RTs at the verb region with overall median FP value in red
df_biondo |> 
  filter(roi == "4") |> 
  mutate(item_median = median(fp, na.rm = T), .by=item) |> 
  arrange(desc(item_median)) |> 
  ggplot()  +
  aes(x = reorder(item, item_median),
      y = fp) +
  labs(title = "By-item first-pass RTs at the verb",
       y = "First-pass reading times (ms)",
       x = "Item ID")+
  geom_hline(
    colour = "red", size = 1,
    aes(yintercept = median(fp, na.rm = T))
  ) +
  geom_boxplot(alpha = 0.5) 
```

### By-participant varying intercepts and slopes

```{r}
#| echo: false

# overall intercept/slopes
sum_overall_mean <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySE(measurevar = "fp", na.rm = T) |> 
  mutate(effect = "intercept",
         level = "intercept") |> 
  relocate(effect, level) |> 
  select(-`.id`)

sum_overall_tense <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("verb_t"), na.rm = T) |> 
  mutate(effect = "verb_t") |> 
  rename(level = verb_t) |> 
  relocate(effect, level)

sum_overall_gramm <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("gramm"), na.rm = T) |> 
  mutate(effect = "gramm") |> 
  rename(level = gramm) |> 
  relocate(effect, level)

sum_overall_effects <- rbind(sum_overall_mean, sum_overall_gramm, sum_overall_tense) |> 
  mutate(
    effect = factor(effect, levels = c("intercept", "verb_t", "gramm")),
    level = factor(level, levels = c("intercept", "Past", "Future", "gramm", "ungramm"))
  ) 


# by-participant intercept/slopes
sum_sj_mean <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("sj"), na.rm = T) |> 
  mutate(effect = "intercept",
         level = "intercept") |> 
  relocate(sj, effect)


sum_sj_tense <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("sj", "verb_t"), na.rm = T) |> 
  mutate(effect = "verb_t") |> 
  rename(level = verb_t) |> 
  relocate(sj, effect)

sum_sj_gramm <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("sj", "gramm"), na.rm = T) |> 
  mutate(effect = "gramm") |> 
  rename(level = gramm) |> 
  relocate(sj, effect)

sum_sj_effects <- rbind(sum_sj_mean, sum_sj_gramm, sum_sj_tense) |> 
  mutate(
    effect = factor(effect, levels = c("intercept", "verb_t", "gramm")),
    level = factor(level, levels = c("intercept", "Past", "Future", "gramm", "ungramm"))
  )

# plot by-sj intercept/slopes with grand mean/effects
fig_sj_effects <-
  sum_sj_effects |> 
  filter(effect != "intercept") |> 
  ggplot() +
  aes(x = level, y = fp, colour = sj) +
  facet_wrap(~effect, scales = "free_x") +
  geom_point(alpha = .4) +
  geom_line(alpha = .4,
      aes(group = sj, colour = sj)) +
  geom_vline(xintercept = 1.5, colour = "grey", linetype = "dashed") +
  theme(legend.position = "none",
        axis.title.x = element_blank()) +
    stat_summary(
    geom = "point",
    fun.y = "mean",
    col = "black",
    size = 2.5,
    shape = "square",
    fill = "black"
  ) +
    stat_summary(
    geom = "line",
    group = 1,
    fun.y = "mean",
    col = "black",
    size = 1
  ) +
  ylim(200,925)+
  labs(title = "By-participant means per factor level")

fig_sj_intercept <-
  sum_sj_effects |> 
  filter(effect == "intercept") |> 
  ggplot() +
  aes(x = reorder(sj,fp), y = fp) +
  facet_wrap(~effect, scales = "free_x") +
  geom_point(alpha = .4, aes(colour = sj))  +
  geom_point(data = sum_overall_effects[sum_overall_effects$effect == "intercept",],
             aes(y = fp), x = 30.5) +
  geom_hline(data = sum_overall_effects[sum_overall_effects$effect == "intercept",],
             aes(yintercept = fp),
             linetype = "dashed",
             alpha = 0.5) +
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.title.x = element_blank())+
  ylim(200,925) +
  labs(title = "By-participant overall means")
  

  
```

```{r}
#| label: fig-pred-fp-verb-bypx
#| fig-cap: Predicted by-participant varying intercepts (A) and slopes (B) with overall effects in black
#| echo: false
#| fig-asp: .5

fig_sj_intercept + fig_sj_effects +
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(.3,.6))
```

### By-item varying intercepts and slopes

```{r}
#| echo: false

# overall intercept/slopes
sum_overall_mean <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySE(measurevar = "fp", na.rm = T) |> 
  mutate(effect = "intercept",
         level = "intercept") |> 
  relocate(effect, level) |> 
  select(-`.id`)

sum_overall_tense <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("verb_t"), na.rm = T) |> 
  mutate(effect = "verb_t") |> 
  rename(level = verb_t) |> 
  relocate(effect, level)

sum_overall_gramm <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("gramm"), na.rm = T) |> 
  mutate(effect = "gramm") |> 
  rename(level = gramm) |> 
  relocate(effect, level)

sum_overall_effects <- rbind(sum_overall_mean, sum_overall_gramm, sum_overall_tense) |> 
  mutate(
    effect = factor(effect, levels = c("intercept", "verb_t", "gramm")),
    level = factor(level, levels = c("intercept", "Past", "Future", "gramm", "ungramm"))
  ) 


# by-item intercept/slopes
sum_item_mean <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("item"), na.rm = T) |> 
  mutate(effect = "intercept",
         level = "intercept") |> 
  relocate(item, effect)


sum_item_tense <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("item", "verb_t"), na.rm = T) |> 
  mutate(effect = "verb_t") |> 
  rename(level = verb_t) |> 
  relocate(item, effect)

sum_item_gramm <-
  df_biondo |>
  filter(roi == "4") |> 
  Rmisc::summarySEwithin(measurevar = "fp", withinvars = c("item", "gramm"), na.rm = T) |> 
  mutate(effect = "gramm") |> 
  rename(level = gramm) |> 
  relocate(item, effect)

sum_item_effects <- rbind(sum_item_mean, sum_item_gramm, sum_item_tense) |> 
  mutate(
    effect = factor(effect, levels = c("intercept", "verb_t", "gramm")),
    level = factor(level, levels = c("intercept", "Past", "Future", "gramm", "ungramm"))
  )

# plot by-item intercept/slopes with grand mean/effects
fig_item_effects <-
  sum_item_effects |> 
  filter(effect != "intercept") |> 
  ggplot() +
  aes(x = level, y = fp, colour = item) +
  facet_wrap(~effect, scales = "free_x") +
  geom_point(alpha = .4) +
  geom_line(alpha = .4,
      aes(group = item, colour = item)) +
  geom_vline(xintercept = 1.5, colour = "grey", linetype = "dashed") +
  theme(legend.position = "none",
        axis.title.x = element_blank()) +
    stat_summary(
    geom = "point",
    fun.y = "mean",
    col = "black",
    size = 2.5,
    shape = "square",
    fill = "black"
  ) +
    stat_summary(
    geom = "line",
    group = 1,
    fun.y = "mean",
    col = "black",
    size = 1
  ) +
  ylim(200,925)+
  labs(title = "By-item means per factor level")

fig_item_intercept <-
  sum_item_effects |> 
  filter(effect == "intercept") |> 
  ggplot() +
  aes(x = reorder(item,fp), y = fp) +
  facet_wrap(~effect, scales = "free_x") +
  geom_point(alpha = .4, aes(colour = item))  +
  geom_point(data = sum_overall_effects[sum_overall_effects$effect == "intercept",],
             aes(y = fp), x = 48.5) +
  geom_hline(data = sum_overall_effects[sum_overall_effects$effect == "intercept",],
             aes(yintercept = fp),
             linetype = "dashed",
             alpha = 0.5) +
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.title.x = element_blank())+
  ylim(200,925) +
  labs(title = "By-item overall means")
  

  
```

```{r}
#| label: fig-fp-verb-byitem
#| fig-cap: By-item varying intercepts (A) and slopes (B) with overall effects in black
#| echo: false
#| fig-asp: .5

fig_item_intercept + fig_item_effects +
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(.3,.6))
```

## Comparing participant and item

- just by eye-balling our data we see there was more variability in intercepts and effects by-participant than by-item
  + this is typical: People tend to vary more than our highly controlled experimental items

- how can we take this variability of both item and participant into account?
  + mixed models!
  + today we'll focus on varying intercepts, first for by-participants and then by-item

# Random intercepts

- random intercepts = taking group-level variance in overall tendencies into account

## Random intercepts: one grouping factor

- below is our first *mixed effects model*

```{r}
fit_lmm_fp_sj <-
  lmer(log(fp) ~ verb_t*gramm + 
         (1|sj), 
       data = df_biondo, 
       subset = roi == 4) 
```
1. create an object `fit_lmm_fp_sj`, which contains...
1. a mixed model (`lmer()`): log first-pass RTs as a function of our fixed effects, plus...
2. varying intercepts (`1`) by-participant (`|sj`)
3. from our dataset
4. subsetted to only include the verb region
    + can also be done above the model using `filter(roi == 4)`

### Summary {.smaller}

- we can use the `summary()` function, just as we did with (`g`)`lm()`

```{r}
#| output-location: column-fragment
summary(fit_lmm_fp_sj)
```

### Model info


```{r}
#| eval: false
#| code-annotations: right
Linear mixed model fit by REML ['lmerMod']. #<1>
Formula: log(fp) ~ verb_t * gramm + (1 | sj) #<2>
   Data: df_biondo #<3>
 Subset: roi == 4 #<4>

REML criterion at convergence: 4479.1 #<5>

Scaled residuals: #<6>
    Min      1Q  Median      3Q     Max 
-4.0560 -0.6427 -0.0419  0.6168  4.0901 
```

1. Model description (object class = `lmerMod`)
2. Model formula
3. Data
4. Any subsetting
5. REML: Restricted Maximum Likelihood; important for model comparison, but not for us today
6. Residuals: do these look normally distributed?

### Fixed effects

```{r}
#| eval: false
Fixed effects: #<1>
                Estimate Std. Error t value #<2>
(Intercept)     5.957102   0.033809 176.199 #<3>
verb_t1         0.062209   0.013787   4.512 #<4>
gramm1          0.003466   0.013787   0.251 #<5>
verb_t1:gramm1 -0.015741   0.027573  -0.571 #<6>

Correlation of Fixed Effects:  #<7>
            (Intr) vrb_t1 gramm1
verb_t1      0.000              
gramm1       0.000 -0.002       
vrb_t1:grm1  0.000  0.002  0.000
```

1. Our fixed effects:
2. Estimate (coefficient), standard error, t-value (no p-value...)
3. Intercept (grand mean)
4. Effect of tense
5. Effect of grammaticality
6. Interaction Effect
7. Correlation matrix of fixed effects

### Random effects

```{r}
#| eval: false
Random effects: #<1>
 Groups   Name        Variance Std.Dev. #<2>
 sj       (Intercept) 0.06573  0.2564  #<3>
 Residual             0.18030  0.4246 #<4>
Number of obs: 3795, groups:  sj, 60 #<5>
```

1. Random effects
2. Grouping factor, effect name, overall variance and standard deviation
3. By-participant random intercepts
4. Residual error not accounted for by our random effects
5. Number of observations and grouping factor levels

## Interpreting random effects

- we can also selectively print our random effects (variance components) using the `VarCorr()` function from `lme4`
  + only gives us the standard deviation, which is the square root of the variance

```{r}
#| output-location: column-fragment
VarCorr(fit_lmm_fp_sj)
```

- to also get the variance

```{r}
#| output-location: column-fragment
print(VarCorr(fit_lmm_fp_sj), 
      comp = c("Variance", "Std.Dev"))
```

- if we compute the mean, variance, and SD of the by-participant intercepts we get

```{r}
#| output-location: column-fragment
#| code-fold: true
coef(fit_lmm_fp_sj) |> 
  pluck("sj") |> 
as_tibble() |> rownames_to_column(var = "sj") |> 
  rename(
    intercept = 2
  ) |> 
  summarise(
    mean = mean(intercept),
    var = var(intercept),
    sd = sd(intercept)
  )
```

### 68-95% rule

- in a normal distribution, 68% of the data will lie within +/-1 SD of the mean, and 95% will lie within +/-2 SDs of the mean
- our model intercept is `r fixef(fit_lmm_fp_sj)[1]`, so 68% of our participant intercepts will lie roughly between `r fixef(fit_lmm_fp_sj)[1]-0.2563809` and `r fixef(fit_lmm_fp_sj)[1]+0.2563809`

::: {.column width="35%"}
```{r}
#| echo: false
print(VarCorr(fit_lmm_fp_sj), 
      comp = c("Variance", "Std.Dev"))
```

- so we can interpret the standard deviation as quantifying the variability of the by-participant intercepts around the mean (i.e., intercept)
:::

::: {.column width="60%"}
```{r}
#| echo: false
sj_int <- coef(fit_lmm_fp_sj) |> 
  pluck(1) |> rename(intercept = `(Intercept)`)

# set quantiles
q5 <- quantile(sj_int$intercept,.025)
q95 <- quantile(sj_int$intercept,.975)
q16 <- quantile(sj_int$intercept,.16)
q84 <- quantile(sj_int$intercept,.84)

x.dens <- density(sj_int$intercept)
df.dens <- data.frame(x = x.dens$x, y = x.dens$y)

# plot

fig_res_density <-
  sj_int |> 
  ggplot() +
  aes(x = intercept) +
  labs(title = "By-participant varying intercepts density",
       x = "Intercept",
       y = "Density") +
  geom_density() +
  geom_area(data = subset(df.dens, x >= q5 & x <= q95), 
              aes(x=x,y=y), fill = 'red',
             alpha = .4) +
  geom_area(data = subset(df.dens, x >= q16 & x <= q84), 
              aes(x=x,y=y), fill = 'blue',
             alpha = .4) +
  geom_vline(colour = "black",
             linetype = "dashed",
             aes(xintercept = fixef(fit_lmm_fp_sj)[1])) +
  geom_vline(colour = "blue",
             linetype = "dashed",
             aes(xintercept = q16)) +
  geom_vline(colour = "blue",
             linetype = "dashed",
             aes(xintercept = q84)) +
  geom_vline(colour = "red",
             linetype = "dashed",
             aes(xintercept = q5)) +
    # add labels
    geom_text(x = q5, y = 1, label = "mean - SD*1.96", angle = 90,
              vjust = -.35,
              colour = "red")+
  geom_vline(colour = "red",
             linetype = "dashed",
             aes(xintercept = q95)) +
    geom_text(x = q95, y = 1, label = "mean + SD*1.96",
              colour = "red", angle = 90,
              vjust = -.35) +
    geom_text(x = q16, y = 1.3, label = "mean - SD",
              colour = "blue", angle = 90,
              vjust = -.35) +
    geom_text(x = q84, y = 1.3, label = "mean + SD",
              colour = "blue", angle = 90,
              vjust = -.35) +
    geom_text(aes(x = mean(intercept)), y = 1, label = "Intercept", angle = 90,
              vjust = -.35) 
```

```{r}
#| label: fig-density
#| fig-cap: Density plot of by-participant intercepts with 95% (+/-SD*1.96) and 68% (+/-SD) ranges
#| echo: false
fig_res_density
```

:::

## `lmerTest::lmer()`

- recall we had a conflict for the function `lmer()` between `lme4` and `lmerTest`, and we set `lme4` to be our default for this function
- let's refit our model using `lmerTest::lmer()`

```{r}
fit_lmm_fp_sj <-
  lmerTest::lmer(log(fp) ~ verb_t*gramm + 
         (1|sj), 
       data = df_biondo, 
       subset = roi == 4) 
```

- the only change is the addition of `lmerTest::`, which specifies which package to retreive the `lmer()` function from

## {.smaller}

- what's different in the summary?

```{r}
#| output-location: column
summary(fit_lmm_fp_sj)
```

##

```{r}
#| eval: false
Linear mixed model fit by REML. t-tests use Satterthwaite's method ['lmerModLmerTest'] #<1>
...

Fixed effects:
                  Estimate  Std. Error          df t value             Pr(>|t|)     #<2>
(Intercept)       5.957102    0.033809   58.991899 176.199 < 0.0000000000000002 ***
verb_t1           0.062209    0.013787 3732.065822   4.512           0.00000662 ***
gramm1            0.003466    0.013787 3732.032139   0.251                0.802    
verb_t1:gramm1   -0.015741    0.027573 3732.037124  -0.571                0.568    

...
```

1. New: `Satterthwaite's method` and object class (`lmerModLmerTest`)
2. Fixed effects include `df` (degrees of freedom) and *p*-values (`Pr(>|t|)`)

- `lme4::lmer()` doesn't provide degrees of freedom or *p*-values
  + defining degrees of freedom (and therefore calculating *p*-values) is more complex and not trivial in mixed models
  + `lmerTest` uses the Satterthwaite method, which is fine for our purposes
- importantly, everything else is exactly the same as when we use `lme4::lmer()`

## Fixed effects for `lm()` and `lmer()`

- let's compare our fixed effects to those from a model without random effects

```{r}
fit_lm_fp <-
  lm(log(fp) ~ verb_t*gramm, 
       data = df_biondo, 
       subset = roi == 4) 
```

### Comparing fixed effects

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lm_fp_sj
#| tbl-cap: Fixed effects for `lm_fp_sj`
tidy(fit_lm_fp) |> 
  as_tibble() |> 
  mutate(p.value = round(p.value,8),
         across(c(estimate, std.error,statistic),round,3)) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lmm_fp_sj
#| tbl-cap: Fixed effects for `lmm_fp_sj`
broom.mixed::tidy(fit_lmm_fp_sj,
                    effects="fixed") |> 
  mutate(p.value = round(p.value,8),
         across(c(estimate, std.error,statistic),round,3),
         df = round(df,0)) |> 
  select(-effect) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::


::: {.column width="100%"}
- so far we see that our model estimates are still descriptively similar, there are only some slight quantitative differences
- your fixed effects will typically be unchanged with the addition of random effects
  + what changes will be usually be the standard error, *t*-value (or *z*-value for generalised linear (mixed) models), confidence intervals, and *p*-values
  + the magnitude of this change will depend on whether the inclusion of the random effects better accounts for variability in your data than your fixed effects alone
  
:::

::::

### Comparing residual error

- the residual error for our fixed-effects-only model was is `r round(glance(fit_lm_fp)$sigma,2)`

```{r}
glance(fit_lm_fp)$sigma
```

- for our by-participant varying intercepts model it goes down to `r round(glance(fit_lmm_fp_sj)$sigma,2)`

```{r}
glance(fit_lmm_fp_sj)$sigma
```

- this tells us that our inclusion of by-participant varying intercepts accounts for some of the variance in the model that was not accounted for in fixed-effects-only model

- are there any other possible sources of variance that we haven't taken into account?

# Crossed random effects: two grouping factors

- we still haven't taken by-item variance into account, let's now include *crossed* random effects in our model

```{r}
fit_lmm_fp_sj_item <-
  lmerTest::lmer(log(fp) ~ verb_t*gramm +
                   (1|sj) +
                   (1|item),
                 data = df_biondo,
                 subset = roi == 4)
```

- crossed random effects refer to a property of your data (/experimental design), i.e., repeated measures for items *and* participants
  + one level of a grouping factor contains observations from all levels from another grouping factor (e.g., each item has an observations from each participant and vice versa)

###

```{r}
#| output-location: column
summary(fit_lmm_fp_sj_item)
```

### Comparing random effects

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lmm_fp_sj_res
#| tbl-cap: By-participant varying variance component
broom.mixed::tidy(fit_lmm_fp_sj,
                    effects="ran_pars") |> 
  select(-effect) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lmm_fp_sj_item_res
#| tbl-cap: By-participant and -item variance components
broom.mixed::tidy(fit_lmm_fp_sj_item,
                    effects="ran_pars") |> 
  select(-effect) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::

::: {.column width="100%"}
- we now have the variance of by-item random intercepts (@tbl-lmm_fp_sj_item_res)
- the *variance component* for the by-subjects intercepts is not much changed
  + the value is slightly different because it now also takes by-item variability into account
- the *residual error* also goes down in `fit_lmm_fp_sj_item` (`r round(glance(fit_lmm_fp_sj_item)$sigma,2)`), because by-item intercepts account for some of the variance that was unaccounted for in `fit_lmm_fp_sj` (`r round(glance(fit_lmm_fp_sj)$sigma,2)`)
- note that there is most by-participant than by-item variance, this is typical and reflects what we saw in our boxplots

:::

::::

## Comparing fixed effects

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lmm_fp_sj2
#| tbl-cap: Fixed effects for `lmm_fp_sj`
broom.mixed::tidy(fit_lmm_fp_sj,
                    effects="fixed") |> 
  mutate(p.value = round(p.value,8),
         across(c(estimate, std.error,statistic),round,3),
         df = round(df,0)) |> 
  select(-effect) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-lmm_fp_sj_item
#| tbl-cap: Fixed effects for `lmm_fp_sj_item`
broom.mixed::tidy(fit_lmm_fp_sj_item,
                    effects="fixed") |> 
  mutate(p.value = round(p.value,8),
         across(c(estimate, std.error,statistic),round,3),
         df = round(df,0)) |> 
  select(-effect) |> 
  kable() |> 
  kable_styling(font_size = 24)
```
:::


::: {.column width="100%"}
- again we see there isn't much change to our coefficient estimates
:::

::::

### Comparing predictions {.smaller}

```{r}
#| label: fig-plot-comparison
#| fig-cap: Comparison of predicted estimates and 95% confidence intervals for the three models
#| code-fold: true
#| code-summary: Code for plots
#| out-width: "100%"
#| fig-asp: .3
sjPlot::plot_model(fit_lm_fp, type = "int") +
  geom_line(position = position_dodge(.1)) +
  ylim(340,440) +
  labs(title = "fit_lm_fp") +

sjPlot::plot_model(fit_lmm_fp_sj, type = "int") +
  geom_line(position = position_dodge(.1)) +
  ylim(340,440) +
  labs(title = "fit_lmm_fp_sj") +

sjPlot::plot_model(fit_lmm_fp_sj_item, type = "int") +
  geom_line(position = position_dodge(.1)) +
  ylim(340,440) +
  labs(title = "fit_lmm_fp_sj_item") +
  
  plot_layout(guides = "collect")
```

- if we plot the results from all three models we've fit so far we see the estimates are similar but the confidence intervals are wider for the mixed models
  + this is despite the fact that the *p*-values are significant for all three
  
### Model comparison {.smaller}

- we can use the `anova()` function to compare model fit

```{r}
anova(fit_lmm_fp_sj, fit_lmm_fp_sj_item)
```

- here we see that the AIC, BIC, and logLik are all lower for our model with by-participant and -item varying intercepts
  + *lower* AIC and BIC indicate better model fit
  + *higher* logLik indicates better fit
- the inclusion of by-item random intercepts significantly improves the fit of our model

# Exploring our random effects estimates

- what we saw in our model summary were the variance components
  + a description of the variance of our by-item and by-participant random intercepts
- our model also contains intercept estimates for each level of item and participant
  + we can extract the intercept estimates
  + or we extract their deviance from the model intercept

## Extracting fixed effects

- we've already used `coef()` to extract fixed effect estimates from `lm` objects

```{r}
#| output-location: column-fragment
coef(fit_lm_fp)
```

- to extract our fixed effect estimates from `lmer` objects we need `fixef()`

```{r}
#| output-location: column-fragment
fixef(fit_lmm_fp_sj_item)
```

- or we can append `$coefficients` to the model summary

```{r}
#| output-location: column-fragment
summary(fit_lmm_fp_sj_item)$coefficients |> 
  as_tibble()
```

### Extract random intercept estimates

- `coef()` behaves very differently with `lmer` objects, extracting the random effects estimates per level

```{r}
coef(fit_lmm_fp_sj_item) |> pluck("item") |> 
  rownames_to_column(var = "item") |> head()
```

- which outputs a `list` object, with one data frame for `item` and one for `sj`
  + in the code above I've 'plucked' just the by-item coefficients

##

- we can extract just one or the other (`head()` is for presentation purposes):

```{r}
#| output-location: column-fragment
coef(fit_lmm_fp_sj_item) |> pluck("item") |> 
  rownames_to_column(var = "item") |> head()
```

```{r}
#| output-location: column-fragment
coef(fit_lmm_fp_sj_item) |> pluck("sj") |> 
  rownames_to_column(var = "sj") |> head()
```

- why do our intercepts vary, but not `verb_t1`, `gramm1`, or `verb_t1:gramm1`?

```{r}
#| echo: false
coef_fp_item_int <-
  coef(fit_lmm_fp_sj_item) %>%
  pluck("item") |> 
  rownames_to_column(var = "item") |> 
  rename(
    intercept = 2
  ) |> 
  as_tibble()

coef_fp_sj_int <-
  coef(fit_lmm_fp_sj_item) %>%
  pluck("sj") |> 
  rownames_to_column(var = "sj") |> 
  rename(
    intercept = 2
  ) |> 
  as_tibble()


```

### Extract deviations from the intercept

- the `ranef()` function provides the deviance from the model intercept and each random intercept estimate
  + the output is a `list` with a one element per grouping factor

```{r}
ranef(fit_lmm_fp_sj_item)
```

## 

:::: {.columns}

::: {.column width="100%"}
- `ranef()$grouping_factor` or `pluck("grouping_factor")` selects the relevant grouping factor
:::

::: {.column width="50%"}
```{r}
#| output-location: fragment
ranef(fit_lmm_fp_sj_item)$sj |> 
  head()
```
:::

::: {.column width="50%"}
```{r}
#| output-location: fragment
ranef(fit_lmm_fp_sj_item) |> 
  pluck("sj") |> head()
```
:::
::::

### Compare estimates and deviances

- the values extracted by `ranef()` (`sj_dev` in @tbl-res-dev) equal the difference (`difference`) between the model intercept (`model_intercept`) and the by-participant random intercept estimates (`sj_est`)
- so we can either look at each participant's (or item's) estimate, or look at how much it deviates from the model intercept

```{r}
#| echo: false
#| label: tbl-res-dev
#| tbl-cap: Random intercept estimates versus deviance
coef_lmm_sj <-
  coef(fit_lmm_fp_sj_item) |> 
  pluck("sj") |> 
  rownames_to_column(var = "sj") |> 
  rename(
         sj_est = 2)

dev_lmm_sj <-
  ranef(fit_lmm_fp_sj_item) |> 
  pluck("sj") |> 
  rownames_to_column(var = "sj") |> 
  rename(
         sj_dev = 2)

left_join(coef_lmm_sj, dev_lmm_sj, by = "sj") |> 
  relocate(sj, sj_est, sj_dev) |> 
  select(sj, sj_est, sj_dev) |> 
  mutate(est_minus_dev = sj_est - sj_dev,
         model_intercept = fixef(fit_lmm_fp_sj_item)[1],
         est_minus_intercept = sj_est - model_intercept) |> 
  # relocate(sj) |> 
  head() |> 
  kable(digits = 3) |> 
  kable_styling(font_size = 24) 
```

## Visualise random effects

:::: {.columns}

::: {.column width="100%"}
- the `lattice` package automatically produces plots of random effects estimates
:::

::: {.column width="50%"}
```{r}
lattice::dotplot(ranef(fit_lmm_fp_sj_item))$item
```
:::

::: {.column width="50%"}
```{r}
lattice::dotplot(ranef(fit_lmm_fp_sj_item))$sj
```
:::

::::

# Reporting your model

- according to @sonderegger_regression_nodate-1 (p. 297), we should report:
  
  1. model definition (sometimes in 'Data Analysis' section)
  2. Fixed effects
  3. Random effects
  4. Sample size (number of observations, number of levels for each grouping factor)
  5. one or more quantitative summaries of the model, e.g., AIC, BIC, or logLik (although they're only informative in comparison to another model fit to the same data)

## Model definition

> We conducted the analysis by fitting `linear mixed-effect models` to our data, using the R package lme4 (Bates et al., 2014). We included Time Reference (past, future), and Verb Match (match, mismatch) as `fixed-effect factors` [...] by adopting `sum contrast coding` (Schad et al., 2020): past and match conditions were `coded as –.5`. while future and mismatch conditions were `coded as .5`. [...] Moreover, we included `crossed random intercepts` and random slopes for all fixed-effect parameters `for subject and item` grouping factors (Barr et al., 2013) in all models. [...] Logit mixed-effect models were employed (Jaeger, 2008) for the analysis of the probability of regression measure. [...] P-values were derived by using the lmerTest package [@lmerTest_package].

--- @biondo_yesterday_2022, p. 9

###

- could also explicitly mention method used for *p*-values, an example:

> P-values for individual predictors were computed using `lmerTest`, with the `Satterthwaite` option for denominator degrees of freedom for F statistics.

--- @troyer_catch_2020, p. 9

- but here they don't cite the package
  + so you see, there's alway something you miss...
- FYI,  to get a package's citation, run `citation("lmerTest")` in the Console

## Results

- a combination of tables, figures, and in-text coefficient estimates is always key
- in-text, the *t*- and *p*-values should be included at minimum, Estimate and standard error (*Est* = ..., *SE* = ...,) could also be included if you aren't reporting many effects but must at least be included in a table
- figures will typically only show the distribution of raw observations and model predictions for fixed effects

### In-text

A main effect of tense was found in first-pass reading times at the verb region (*Est* = 0.062, *t* = 4.8, *p* < .001), with the future tense (*M* = `r round(mean(df_biondo$fp[df_biondo$roi=="4" & df_biondo$verb_t=="Future"], na.rm=T),0)`ms, *SD* = `r round(sd(df_biondo$fp[df_biondo$roi=="4" & df_biondo$verb_t=="Future"], na.rm=T),0)`ms) eliciting longer first-pass reading times than the past tense.

### Tables

#### Fixed effects

```{r}
#| label: tbl-report-fixed
#| tbl-cap: Table of fixed effects from fit_lmm_fp_sj_item
#| code-fold: true
#| code-summary: Code for table
tidy(fit_lmm_fp_sj_item,
     effects = "fixed") |> 
  as_tibble() |> 
  select(-effect) |> 
  mutate(p.value = format_pval(p.value),
         across(c(estimate,std.error, statistic), round, 3),
         df = round(df,1)) |> 
  mutate(term = fct_recode(term,
    "Intercept" = "(Intercept)",
    "Tense" = "verb_t1",
    "Grammticality" = "gramm1",
    "Tense x Gramm" = "verb_t1:gramm1"
  )) |> 
  kable(
        col.names = c("Coefficient", "$\\hat{\\beta}$", "SE", "t", "df", "p")) |> 
  kable_styling()
```

#### Random effects

```{r}
#| label: tbl-report-res
#| tbl-cap: Table of random effects from fit_lmm_fp_sj_item
#| code-fold: true
#| code-summary: Code for table
as.data.frame(VarCorr(fit_lmm_fp_sj_item),comp=c("Variance","Std.Dev.")) |> 
  as_tibble() |> 
  select(-var2) |> 
  # mutate(var1 = ifelse(var1 == "NA", " ", var1)) |>
  kable(digits = 3,
        col.names = c("Group", "Term", "Variance", "SD")) |> 
  kable_styling()
```

### Figures

- we don't usually include plots of our random effects in publications, but these can be useful for model exploration and can be included in supplementary materials

#### `lattice`

- as already mentioned, we can simply use the the `lattice` package

```{r}
#| label: fig-lattice
#| fig-cap: By-participant varying slopes (`lattice::dotplot(res(model))`)
library(lattice)

dotplot(ranef(fit_lmm_fp_sj_item))["sj"]
```

#### `broom.mixed` {.smaller}

- or we can also generate the same plots using `tidy()` from the `broom.mixed` package + `ggplot()` (@fig-report A)
- and we can add the model intercept to get each by-participant estimate, i.e., the values we get with `coef()`  (@fig-report B)

```{r}
#| label: fig-report
#| fig-cap: Back-transformed first-pass reading times (ms) at the verb region with 95% CIs
#| code-fold: true
#| code-summary: Code for plot
#| out-width: "100%"
#| fig-asp: .4

fig_res_dev <-
  broom.mixed::tidy(fit_lmm_fp_sj_item, effects = "ran_vals", conf.int = TRUE) |> 
  filter(group == "sj") |> 
  ggplot() +
  aes(x = estimate, y = reorder(level, estimate))  +
  labs(title = "By-participant varying intercepts with 95% CIs",
       y = "Participant ID",
       x = "Deviation from the intercept") +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed") +
  geom_point(colour = "blue") +
  geom_errorbar(
    aes(xmin = conf.low,
        xmax = conf.high)
  ) +
  scale_x_continuous(breaks = c(-0.5,0,0.5)) +
  facet_grid(~term)

fig_res_est <- broom.mixed::tidy(fit_lmm_fp_sj_item, effects = "ran_vals", conf.int = TRUE) |> 
  filter(group == "sj") |> 
  # back-transform to ms
  mutate(across(c(estimate,conf.low,conf.high),~.+fixef(fit_lmm_fp_sj_item)[1])) |> 
  # mutate(across(c(estimate,conf.low,conf.high),exp)) |> 
  # plot
  ggplot() +
  aes(x = estimate, y = reorder(level, estimate))  +
  labs(title = "By-participant varying intercepts with 95% CIs",
       y = "Participant ID",
       x = "Estimate (log)") +
  geom_vline(xintercept = fixef(fit_lmm_fp_sj)[1], colour = "red", linetype = "dashed") +
  geom_point(colour = "blue") +
  geom_errorbar(
    aes(xmin = conf.low,
        xmax = conf.high)
  ) +
  scale_x_continuous(breaks = c(5.457102,5.957102 ,6.457102)) +
  facet_grid(~term)

fig_res_dev + fig_res_est +
  plot_annotation(tag_levels = "A")
```

### {.smaller}

- and we can back-transform these values to milliseconds by exponentiating the estimates in the log scale (@fig-report-ms A)
- and we can back-transform deviances by subtracting the exponentiating model estimate from the back-transformed estimates (@fig-report-ms B)

```{r}
#| label: fig-report-ms
#| fig-cap: By-participant estimates back-transformed to milliseconds
#| code-fold: true
#| code-summary: Code for plot
#| out-width: "100%"
#| fig-asp: .4
fig_res_est_ms <-
  broom.mixed::tidy(fit_lmm_fp_sj_item, effects = "ran_vals", conf.int = TRUE) |> 
  filter(group == "sj") |> 
  # back-transform to ms
  mutate(across(c(estimate,conf.low,conf.high),~.+fixef(fit_lmm_fp_sj_item)[1])) |> 
  mutate(across(c(estimate,conf.low,conf.high),exp)) |>
  # plot
  ggplot() +
  aes(x = estimate, y = reorder(level, estimate))  +
  labs(title = "By-participant varying intercepts with 95% CIs",
       y = "Participant ID",
       x = "Estimate (ms)") +
  geom_vline(xintercept = exp(fixef(fit_lmm_fp_sj)[1]), colour = "red", linetype = "dashed") +
  geom_point(colour = "blue") +
  geom_errorbar(
    aes(xmin = conf.low,
        xmax = conf.high)
  ) +
  scale_x_continuous(breaks = c(186.4884,386.4884, 586.4884, 786.4884)) +
  facet_grid(~term)

fig_res_dev_ms <-
broom.mixed::tidy(fit_lmm_fp_sj_item, effects = "ran_vals", conf.int = TRUE) |> 
  filter(group == "sj") |> 
  # back-transform to ms
  mutate(across(c(estimate,conf.low,conf.high),~.+fixef(fit_lmm_fp_sj_item)[1])) |> 
  mutate(across(c(estimate,conf.low,conf.high),exp)) |>
  mutate(across(c(estimate,conf.low,conf.high),~.-exp(fixef(fit_lmm_fp_sj_item)[1]))) |>
  # plot
  ggplot() +
  aes(x = estimate, y = reorder(level, estimate))  +
  labs(title = "Deviances in by-participant varying intercepts with 95% CIs",
       y = "Participant ID",
       x = "Deviance (ms)") +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed") +
  geom_point(colour = "blue") +
  geom_errorbar(
    aes(xmin = conf.low,
        xmax = conf.high)
  ) +
  # scale_x_continuous(breaks = c(-0.5,0,0.5)) +
  facet_grid(~term)

fig_res_est_ms + fig_res_dev_ms + plot_annotation(tag_levels = "A")
```

# Learning objectives 🏁 {.unnumbered .unlisted .uncounted}

Today we learned...

- what linear mixed models are ✅
- how to fit a random-intercepts model ✅
- how to inspect and interpret a mixed effects model ✅

# Important terms {.unnumbered .smaller .uncounted}

```{r}
#| echo: false
content <- 
  googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/17CqdxKL9lyy-PbTB2ZnfWNWs4oV--CcBvrqlh_aEPGQ/edit?usp=sharing")

content |> 
  filter(`Lecture topic` == "08 - LMMs 1: random intercepts") |> 
  select(-`Lecture topic`) |> 
  gt() 
```

# Task

1. Fit a linear mixed model (`lm()` function) to log-transformed total reading times (`tt`) at the adverb region (`roi == 2`), with adverb time reference (`adv_t`) and gramm (`gramm`) and their interaction as fixed effects and by-participant and by-item varying intercepts. Use sum contrast coding (`Past` and `gramm` = `-0.5`, `Future` and `ungramm` = `+0.5`). Save this model as `fit_lmm_adv_tt`.

```{r}
#| echo: false
#| eval: true

contrasts(df_biondo$adv_t) <- c(-0.5,0.5)

fit_lmm_adv_tt <-
  lmerTest::lmer(
    log(tt) ~ adv_t*gramm +
      (1|sj) +
      (1|item),
    data = df_biondo,
    subset = roi == 2
  )
```


3. Inspect the fixed effect of your model.
```{r}
#| eval: false
#| echo: false
summary(fit_lmm_adv_tt)
```

5. Plot the fixed effects for `fit_lm_adv_tt` and `fit_lmm_adv_tt`.

```{r}
#| eval: false
#| echo: false

ggeffect(fit_lmm_adv_tt,
          effects = "fixed",
          terms = c("adv_t", "gramm")) |> 
  as_tibble() |> 
    # mutate(across(c(estimate,conf.low,conf.high),exp))
  mutate(across(c(predicted,conf.low,conf.high),exp))
```

```{r}
coef_fixed <- 
  broom.mixed::tidy(
  fit_lmm_adv_tt,
  effects="fixed",
  conf.int = T
)

pred_back <-
  tibble(
  tense = c(rep("Past",2),rep("Future",2)),
  gramm = rep(c("gramm","ungramm"),2)
  ) 
```



4. Inspect the random effects for `fit_lmm_adv_tt`. Describe what you see.

6. Plot the random effects per participant and item.

5. Write up a description of your model as if for a publication (model formula, contrasts, random effects structure, packages/methods used). 

6. Write up the results (coefficient estimates, etc.).

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


