---
title: "Independence assumption"
subtitle: "Grouping factors and generalizability"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-01-12
format: 
  pdf:
    output-file: 07-independence_assumption.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    include-in-header: ../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    output-location: fragment
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
csl: ../apa.csl
execute:
  echo: false
  include: false
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn about...

- the independence assumption
- types of  non-independence in linguistic data
- the history of mixed models in linguistics

# Resources {.unnumbered .unlisted}

- this lecture covers
  + Sections 1 and 2 from @winter_independence_2021
  + Sections 14.1-14.3 from Ch. 14 (Mixed Models I) in @winter_statistics_2019
  + Sections 8.1-8.2 from Ch. 8 (Mixed-effects models I: Linear Regression) in @sonderegger_regression_nodate-1

```{r}
# suppress scientific notation
options(scipen=999)
options(pillar.sigfig = 5)
```

```{r}
# load libraries
pacman::p_load(
               tidyverse,
               here,
               broom,
               janitor,
               languageR)
```

```{r}
#| echo: false

# extra packages for the lecture notes/slides
pacman::p_load(
               patchwork,
               knitr,
               kableExtra)
```

```{r}
# set preferred ggplot2 theme
theme_set(theme_bw() + theme(plot.title = element_text(size = 10)))
```

# Independence assumption

- we already learned about some model assumptions
  - assumption of *normality* of residuals
  - *homoscedasticity* (constant variance) of residuals
  - absence of *collinearity* of predictors

- there another, argulably more important assumption
  - assumption of *independence*
  
## (Non-)Independence

- non-independence: any possible link or connection between groups of data points
  + e.g., two observations from the same participant will tend to be more similar than to completely independent observations
  + any case where you might expect some clustering of observations by some grouping factor
- the independence assumption assumes that our data points are *not* linked
  + i.e., the value of one observation is completely independent from another
- violations of this assumption have major implications for Type I (alpha) error
  + i.e., the chances of observing an effect where there is none (false positive)
- it also artificially inflates sample size, which affects statistical power

## Repeated measures design

- the reason most (experimental) linguistic data is non-independent is the use of the repeated-measures design
  + collecting multiple data points from e.g., the same participant and for the same item
  + increases statistical power, needing fewer participants (more data points, lower variance due to control in variability between subjects)
  + saves resources (fewer subjects)

## Other sources of non-independence

- non-independence is prevalent in other fields of linguistics, e.g., 
  + corpus studies: text, author, language, dialect, register
  + phonetic experiments: speaker, listener, exact repetitions
  + socio-phonetics: dialect/geographical proximity, register, speaker

# Pseudoreplication

> Pseudoreplication refers to the treatment of dependent observations as independent data points, which causes an overabundance of erroneously significant results.

--- @winter_pseudoreplication_2011, p. 2137

- analysing nonindependent data as if they were independent
- essentially, violating the independence assumption
  + very (*very*) common in older publications
- can also result in Type M (magnitude) and S (sign) error
- is one contributor (out of many) to the so-called replication crisis

## Problem: Generalizability

- beyond spurious results, how researchers interpret the implications of their findings is problematic

> Unfortunately, outside of a few domains such as psycholinguistics, it remains rare to see psychologists model stimuli as random effects – despite the fact that most inferences researchers draw are clearly meant to generalize over populations of stimuli.

--- @yarkoni_generalizability_2022, p. 4

- if we don't include grouping factors in our models, our findings are not generalisable beyond our sample
  + it could be that our findings are due to a few participants or experimental items who deviate from the rest
- we need to take this by-grouping factor variation into account, but how?

## Solution 1: Averaging

- e.g., repeated measures ANOVA
  + seperate models for by-participant and by-item variance (with averaging) interpreted together

- PRO: takes both by-participant and -item variance into account
- CONs: not flexible or approrpriate for complex designs, and:
  + loses information regarding the variation across the grouped observations
  + lowers N
    + e.g., if we average over participants, we'd have 1 only data point per participant!
  + therefore loses statistical power (Type II error)
  + inflates Type I error (chance of a false positive)

- in sum: not optimal

## Solution 2: Single observations

- run an experiment without repeated measures
  + but this lowers statistical power
  + and drastically reduces generalizability
- e.g., we could present 60 participants with a single item
  + or we could present 1 participant with 60 trials
  + but these findings also can't be generalised beyond that one item or one participant...

- in sum: not optimal

## Solution 3: Linear mixed models

- best available solution: use repeated-measures design and mixed models

- a.k.a. mixed (effects) models/LM(E)Ms, multi-level models, hierarchical models
- "mixed" because they contain:
  + **fixed effects**: usually predictors; describe systematic variation in our data that we wish to explain
  + **random effects**: unsystematic variation that are due to random sampling
- random effects take dependence between observations into account
  + contain varying intercepts and slopes per level of a **grouping factor**
- fixed effects estimates are usually qualititatively unchanged
  + what is affected in the measures of *variance*

# History of mixed-effects models

## 1973: Language-as-fixed-effect-fallacy

- none of these ideas are new to linguistics
- @clark_language-as-fixed-effect_1973:
  + without including dependencies between repeated observations from the same **linguistic items** in our models, we cannot generalise our findings beyond our stimuli
  + our results are relevant only for the subset of the population from which we sampled

###

> The remedies for the language-as-fixed-effect fallacy are for the most part obvious. They include doing the right statistics, choosing the appropriate experimental design, and selecting a random or representative sample of language.

--- @clark_language-as-fixed-effect_1973, p. 347

## ANOVAs: aggregation

- repeated measures ANOVAs were commonly used to take dependence between observations into account (and are still common in come fields today)
  + require aggregation (i.e., averaging) over items *or* subjects, not both simultaneously
  + drastically reduces our number of observations
  + loss of information in the variance of observed data
  + i.e., a loss of power (Type II error) and inflated Type I error (false positive)

## 2008: @baayen_mixed-effects_2008 and `lme4`

- enter mixed models with *crossed* random effects

- Journal of Memory and Language, Special Issue: Emerging Data Analysis
  + @baayen_mixed-effects_2008: introduction of `lme4` package for linear mixed models
  + @jaeger_categorical_2008: overview of generalised linear mixed models
- in addition, @baayen_analyzing_2008 was published, a textbook for analysing linguistic data with R with an emphasis on LMMs with `lme4`


# Learning Objectives 🏁 {.unnumbered .unlisted}

Today we learned about...

- the independence assumption ✅
- types of  non-independence in linguistic data ✅
- the history of mixed models in linguistics ✅

# Task

Discuss the following questions.

1. What is the independence assumption?
2. What happens when the independence assumption is violated?
2. What is the language-as-fixed-effect-fallacy?
3. What other sources of variance might be present in language research?
4. Why are repeated measures ANOVAs sub-optimal?

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


